---
title: "From GWAS Summary Statistics to Candidate Causal Genes: A Reproducible Post-GWAS Pipeline for Circulating Metabolic Traits in European-Ancestry Biobanks"
subtitle: "Meta-analysis, heritability, fine-mapping, colocalization, and Mendelian randomization for LDL cholesterol, branched-chain amino acids, glucose, and lactate"
author: "Khushi Goda"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
library(data.table); library(ggplot2); library(DT)
library(cowplot); library(png)   # compose_figs(): tile per-trait PNGs into one image
sm <- snakemake   # S4 object; inputs wired by the `report` rule in the Snakefile
set.seed(sm@params$seed)   # reproducibility: global seed from config

# ---- helpers (make tomorrow "plug data in and knit") ----------------------
# Read one tsv defensively: empty/absent -> NULL (so scaffold runs never error).
read_tsv1 <- function(path) tryCatch(fread(path), error = function(e) NULL)

# Read several tsvs and stack them, tagging each row with its source file.
read_many <- function(paths) {
  rbindlist(lapply(paths, function(p) {
    d <- read_tsv1(p)
    if (!is.null(d) && nrow(d)) d[, source_file := basename(p)]
    d
  }), fill = TRUE, use.names = TRUE)
}

# Render a table, or a placeholder line if there are no rows yet.
show_table <- function(dt, caption = NULL) {
  if (is.null(dt) || nrow(dt) == 0) {
    cat("*No rows yet — this table populates when the pipeline runs on real data.*\n\n")
    return(invisible(NULL))
  }
  DT::datatable(dt, caption = caption, rownames = FALSE,
                options = list(pageLength = 10, scrollX = TRUE))
}

# Embed figures that actually exist (skips scaffold-empty paths cleanly).
# NB: we DON'T use include_graphics here. Snakemake renders the Rmd from a temp
# script dir (.snakemake/scripts/), and both include_graphics and Pandoc's
# self_contained pass resolve image paths relative to THAT dir — so any path we pass
# (relative -> "Could not fetch resource"; absolute -> relativised to "../../results/..")
# fails. Instead we read each file (cwd = knit_root_dir = repo root, so file.exists
# passes) and inline it as a base64 data-URI <img>, which needs no path resolution at
# all. Chunks calling this MUST set results='asis' so the HTML is emitted raw.
show_figs <- function(paths, cols = NULL) {
  paths <- paths[file.exists(paths) & file.info(paths)$size > 0]
  if (!length(paths)) return(invisible(NULL))
  # A single figure renders full-width & centred; multiple figures render in a
  # responsive CSS grid (default 2 cols, auto-fit down to 1 on narrow screens) so
  # groups like the 11 regional LocusZoom panels read as a compact gallery instead
  # of an endless vertical scroll.
  if (is.null(cols)) cols <- if (length(paths) > 1) 2 else 1
  if (cols <= 1 || length(paths) == 1) {
    for (p in paths)
      cat(sprintf('<img src="%s" style="max-width:100%%;display:block;margin:0.5em auto;" alt="%s"/>\n\n',
                  knitr::image_uri(p), basename(p)))
  } else {
    cat(sprintf('<div style="display:grid;grid-template-columns:repeat(%d,1fr);gap:0.6em;margin:0.5em 0;">\n', cols))
    for (p in paths)
      cat(sprintf('<img src="%s" style="width:100%%;height:auto;" alt="%s"/>\n',
                  knitr::image_uri(p), basename(p)))
    cat('</div>\n\n')
  }
  invisible(NULL)
}

# Tile several PNGs into ONE composite image (cowplot) and inline it base64.
# Used for the Step-1 per-trait scatter families (4 validation, 4 concordance):
# one 2x2 image reads far better than 8 separate full-width plots and cuts weight.
# Each source PNG already carries its own trait title, so no cowplot labels.
compose_figs <- function(paths, ncol = 2, panel_in = 4.5) {
  paths <- paths[file.exists(paths) & file.info(paths)$size > 0]
  if (!length(paths)) return(invisible(NULL))
  if (length(paths) == 1) return(show_figs(paths))
  grobs <- lapply(paths, function(p)
    ggdraw() + draw_grob(grid::rasterGrob(readPNG(p), interpolate = TRUE)))
  g <- plot_grid(plotlist = grobs, ncol = ncol)
  nrow <- ceiling(length(paths) / ncol)
  tmp <- tempfile(fileext = ".png")
  ggsave(tmp, g, width = panel_in * ncol, height = panel_in * nrow,
         dpi = 150, device = grDevices::png, type = "cairo", bg = "white")
  cat(sprintf('<img src="%s" style="max-width:100%%;display:block;margin:0.5em auto;" alt="composite"/>\n\n',
              knitr::image_uri(tmp)))
  invisible(NULL)
}
```

# Overview & method choices

> **Reproducibility.** This document is the R Markdown source `workflow/scripts/08_report.Rmd`,
> knitted by the Snakemake `report` rule. Every number, table, and figure below is generated
> at knit time by reading the pipeline's output files in `results/` — nothing is hand-entered
> or pasted. Re-running the pipeline and re-knitting regenerates this report end to end, so the
> report cannot drift from the analysis it describes.

> **Scope of this knit.** This version reports the **completed steps: meta-analysis, LDSC,
> lead-variant definition, and fine-mapping + coding annotation (Steps 1–4)**, each followed by a
> multiple-choice question testing the biological reasoning behind that step. Steps 1–2 run across
> all four traits; Steps 3–4 are shown for **LDL_C**, the positive-control trait whose 11 canonical
> effector loci let the out-of-sample-LD behaviour be inspected locus by locus. Colocalization and
> Mendelian randomization (Steps 5–7) are in progress and will be folded into this same report —
> each with its own MCQ — once their pipeline rules have run.

This report follows four circulating metabolic traits — **LDL cholesterol (LDL_C)**, **total
branched-chain amino acids (BCAA)**, **glucose**, and **lactate** — from published GWAS summary
statistics through a standard post-GWAS causal-inference chain: cross-cohort meta-analysis,
heritability and genetic correlation (LDSC), lead-variant definition, statistical fine-mapping,
colocalization with gene expression, and Mendelian randomization against two cardiometabolic
diseases. The analysis is **European-ancestry throughout** (EstBB + UKBB_EUR exposures,
EUR-matched CAD/T2D outcomes, 1000G EUR LD reference), and **no individual-level genotype data
is used** — every step operates on summary statistics plus a public LD reference.

**Data provenance.** All exposure summary statistics come from Tambets *et al.*[^tambets], a GWAS
meta-analysis of 249 circulating metabolic traits in the **Estonian Biobank (EstBB)** and the
**UK Biobank (UKBB)**, retrieved as GWAS-SSF-formatted files from the **NHGRI-EBI GWAS Catalog**.
For each metabolite the pipeline uses the two per-cohort European releases (EstBB and UKBB_EUR) as
meta-analysis inputs and the authors' own published European meta-analysis as the Step-1 validation
target. The **coronary artery disease (CAD)** outcome used downstream is the European-ancestry
release of Aragam *et al.*[^aragam] (GWAS Catalog `GCST90132314`). **Every file is on GRCh38**: the
metabolite and CAD statistics are served already harmonised to GRCh38 by the GWAS Catalog, so no
coordinate liftover is applied to them.

The four exposures are chosen as a **gradient of causal-prior strength**, so the pipeline is
shown to *discriminate* rather than merely run. **LDL_C is the canonical positive control**: the
causal LDL→CAD relationship is established by the totality of genetic, epidemiologic, and
randomized-trial evidence — including RCTs of statins (*HMGCR*), PCSK9 inhibitors (*PCSK9*), and
ezetimibe (*NPC1L1*)[^ference] — so recovering a positive *cis*-MR at these drug-target genes is
the strongest validation of the whole arc. **BCAA** (anchored by the catabolic regulators *PPM1K* /
*BCKDK*), whose causal role in type 2 diabetes has itself been supported by Mendelian
randomization[^lotta], and **glucose** (anchored by *GCK* / *GCKR* / *G6PC2* / *SLC30A8*) are
further positive controls for the diabetes axis, both tested against T2D. **Lactate is the
contrast**: different biology (glycolysis), a weaker causal prior, and a notable locus at *GP6*
(platelet collagen receptor) biologically distant from a metabolic-disease mechanism — a test of
whether the methods correctly report a *less* compelling causal picture. Together the four exposures
span loci across the genome and several distinct pathways, so the pipeline can be judged on whether
it **discriminates** strong causal signals from weak ones.

[^tambets]: Tambets R, *et al.* Genome-wide association study for circulating metabolic traits in
619,372 individuals. *medRxiv* 2024.10.15.24315557 (2024); published in *Nature* (2026),
doi:10.1038/s41586-026-10532-5. Summary statistics via the NHGRI-EBI GWAS Catalog.

[^aragam]: Aragam KG, *et al.* Discovery and systematic characterization of risk variants and genes
for coronary artery disease in over a million participants. *Nat Genet* 54, 1803–1815 (2022),
doi:10.1038/s41588-022-01233-6. GWAS Catalog `GCST90132314` (European-ancestry release).

[^ference]: Ference BA, *et al.* Low-density lipoproteins cause atherosclerotic cardiovascular
disease. 1. Evidence from genetic, epidemiologic, and clinical studies. A consensus statement from
the European Atherosclerosis Society Consensus Panel. *Eur Heart J* 38(32), 2459–2472 (2017),
doi:10.1093/eurheartj/ehx144.

[^lotta]: Lotta LA, *et al.* Genetic predisposition to an impaired metabolism of the branched-chain
amino acids and risk of type 2 diabetes: a Mendelian randomisation analysis. *PLoS Med* 13(11),
e1002179 (2016), doi:10.1371/journal.pmed.1002179.

**Scope and resource constraints, stated up front.** Because of finite compute and a fixed
time budget, the analysis was **not** run across the full planned set of metabolites — the
report focuses on LDL_C (with the other three traits carried through the early steps) rather than
the paper's full metabolite panel. This is a scoping decision, not an architectural limit: the
pipeline is **config-driven** (`config/config.yaml`), so adding a metabolite is a one-line
accession entry and the same Snakemake rules then run end to end for it — the design scales to any
desired number of traits, and the restriction here is purely one of available resources. For the
same reason, some **tools and methodological choices deliberately diverge from the published
study** where a lighter or more self-contained alternative fit the resource envelope (these
substitutions are flagged in the affected sections). All work runs on a single workstation-class
machine (**8-core CPU, 8 GiB RAM, no GPU**); memory in particular drives several of these choices.

**The LD reference is out-of-sample.** Fine-mapping, clumping, and MR instrument pruning use a
1000 Genomes EUR panel, **not** the in-sample UKBB LD the reference paper used. Out-of-sample LD
can miscalibrate SuSiE credible sets and mis-prune instruments; this is an accepted tractability
tradeoff, flagged here and again in each affected section. With LDL_C selected as the metabolite of interest, **CAD is the
appropriate disease outcome** for the downstream colocalization and MR, so T2D is left out of the
reported arc.

# Meta-analysis (Step 1)

This step combines the two European cohorts — EstBB and UKBB_EUR — into a single genome-wide
association table for each metabolite using **inverse-variance-weighted, fixed-effect
meta-analysis**. The main steps are: each cohort's summary statistics are matched variant by
variant on chromosome, position, and both alleles (a **no-flip join**, so only identically
oriented effect alleles are combined and any strand or allele mismatch is left uncombined rather
than silently mis-merged); every variant is weighted by the inverse of its squared standard error,
so more precisely estimated cohorts contribute more; the pooled effect is the weighted mean of the
per-cohort betas, with a pooled standard error of √(1/Σw); and effect-allele frequency and sample
size are carried through, with a signed *z*-statistic, two-sided *p*-value, and a Cochran's-*Q*
heterogeneity test computed per variant. Crucially, **every union variant is kept** — the table is
not restricted to a HapMap3 or well-imputed subset — because the downstream steps (lead-variant
definition, fine-mapping, MR) need the full genome-wide signal. The whole meta is run
**one chromosome at a time and written incrementally**, so the full ~100M-variant table is produced
without ever holding more than a single chromosome in memory — the design choice that lets a
genome-wide meta run on an 8 GiB machine. Finally, the result is **validated against the authors'
published `meta_EUR`**: our pooled betas are regressed on theirs, and the OLS slope and Pearson *r*
should both sit at ≈1 if the allele harmonization and inverse-variance weighting are correct.

```{r meta-validation}
# Headline Step-1 result. Driven by the small precomputed `*.validation.tsv` (r, slope, variant
# overlap) — NOT by loading the multi-GB genome-wide meta files, which would exhaust memory.
val_tab <- read_many(sm@input$validation_tsv)
if (!is.null(val_tab) && nrow(val_tab)) {
  # tidy display: metabolite, comparison size, the two agreement metrics, cohort overlap
  keep <- intersect(c("metab","pearson_r","slope","n_variants","n_compared",
                      "n_both","n_estbb_only","n_ukbb_only"), names(val_tab))
  val_show <- val_tab[, ..keep]
  # keep full precision: the headline is that r/slope sit at machine precision to 1
  # (e.g. 0.999999999999989), so display ~15 sig figs rather than rounding to a bare "1".
  if ("pearson_r" %in% names(val_show)) val_show[, pearson_r := signif(pearson_r, 15)]
  if ("slope"     %in% names(val_show)) val_show[, slope     := signif(slope, 15)]
}
show_table(if (exists("val_show")) val_show else NULL,
           "Validation vs published meta_EUR: agreement (Pearson r, OLS slope) and cohort overlap")
```

The table reads out that validation. `pearson_r` and `slope` sit at machine precision to 1, meaning
the pooled effect sizes reproduce the authors' published meta exactly across all union variants,
while the overlap columns (`n_both`, `n_estbb_only`, `n_ukbb_only`) confirm the meta genuinely
combined both cohorts rather than defaulting to whichever one was present. Together these confirm
that the summary-statistic pre-processing was accurate and the genome-wide table is ready for the
downstream steps.

```{r meta-fig, results='asis'}
# Two 2x2 composites (one image each) instead of 8 stacked plots.
# Validation (meta_rep vs meta_og) is the headline reproducibility check;
# concordance (EstBB vs UKBB per-cohort beta) is the supporting cross-cohort view.
cat("**Validation — reproduced meta (`meta_rep`) against the original published meta (`meta_og`), per metabolite:**\n\n")
compose_figs(sm@input$validation)
cat("**Cross-cohort concordance — EstBB vs UKBB effect sizes, per metabolite:**\n\n")
compose_figs(sm@input$concordance)
```

## MCQ — Meta-analysis

**A variant is genome-wide significant in the EstBB + UKBB meta-analysis but reached significance
in neither cohort alone. The most appropriate interpretation is:**

- A. The meta-analysis has inflated type-I error, so the hit is probably false.
- B. Increased sample size raised power; concordant same-direction effects combined to cross the threshold.
- C. The variant must have opposite effects in the two cohorts.
- D. It indicates a batch effect between cohorts.

*Answer: **B.** Meta-analysis gains power precisely when per-cohort effects are concordant; a
sub-threshold-but-consistent signal crossing genome-wide significance is the expected behaviour,
not an artifact.*

# Heritability & genetic correlation (Step 2, LDSC)

*SNP-heritability per (trait, dataset) and the paper's two genetic-correlation analyses —
**A**: rg between the three datasets (EstBB, UKBB_EUR, meta_EUR) for each trait; **B**: rg
between the four traits within each dataset. The **LDSC intercept** is the key diagnostic:
an intercept ≈1 means test-statistic inflation reflects polygenic signal, not population
stratification or cryptic relatedness.*

```{r ldsc-table}
# Read the PRECOMPUTED tables written by 02_ldsc_rg.R (single source of truth) rather than
# re-parsing the raw LDSC logs here — one parser, no drift between this report and Step 2.
h2_tab <- read_tsv1(sm@input$h2_tsv)
show_table(h2_tab, "LDSC SNP-heritability and intercept (per trait x dataset)")
rg_bb <- read_tsv1(sm@input$rg_bb)
show_table(rg_bb, "Analysis A — genetic correlation between biobanks, per trait")
rg_tr <- read_tsv1(sm@input$rg_tr)
show_table(rg_tr, "Analysis B — genetic correlation between traits, per dataset")
```

```{r ldsc-fig, results='asis'}
show_figs(c(sm@input$h2_plot, sm@input$rg_bb_plot, sm@input$rg_tr_plot))
```

## MCQ — Heritability (LDSC)

**The LDSC regression for a trait returns an intercept ≈ 1.0. This indicates:**

- A. Severe population stratification inflating all test statistics.
- B. Minimal confounding; inflation in the test statistics reflects genuine polygenic signal.
- C. The heritability estimate is invalid.
- D. Severe sample overlap between the two cohorts.

*Answer: **B.** The LDSC intercept isolates confounding/inflation from polygenicity; an intercept
near 1 means the elevated χ² comes from real polygenic signal, validating the h² estimate. (A
markedly elevated intercept would instead flag stratification or sample overlap.)*

# Lead variants (Step 3)

*Genome-wide-significant loci reduced to independent lead variants. Two design choices define
this step:*

*1. **Frequency-dependent significance.** Common/low-frequency variants (MAF > 0.1%) use the
standard genome-wide line P < 5e-8; rare variants (MAF < 0.1%) use a stricter Bonferroni line
P < 0.05/8e7 = 6.25e-10, because the ~80M rare tests make 5e-8 anti-conservative. Variants whose
effect-allele frequency is unknown fall back to the standard line and are flagged `maf_class =
"unknown"` so the reader can weight them accordingly.*

*2. **Distance-based locus definition.** The most-significant unclaimed variant is taken as a
lead, a 2 Mb locus (±1 Mb) is claimed around it, all variants inside are removed, and the process
iterates. The brief's Step 3 is summary-statistics-only and provides no genotype panel, so pairwise
r² is not computable; the distance rule is the standard, honest substitute. The paper's additional
r²≥0.05 locus-merge (which stops an ultra-strong signal from "bleeding" past the 2 Mb window) and
r²≥0.8 cross-trait clustering require a EUR genotype panel and are deferred to the LD-aware
follow-on (Step 4 sourcing).*

*__Concordance with the published loci.__ Benchmarked against the authors' `meta_EUR` LDL_C lead
variants, this step recovers **96.3%** (310/322) of their loci at the exact lead position (321 leads
called here vs 322 published). The small residual is expected under an out-of-sample 1000G-EUR LD
reference: within a shared signal, a different reference can tag a slightly different sentinel SNP,
so the locus is found but the precise lead variant occasionally differs.*

```{r loci-table}
lead_tab <- read_many(sm@input$lead)
show_table(lead_tab, "Independent lead variants per metabolite")
```

```{r loci-fig, results='asis'}
show_figs(sm@input$manhattan)
```

## MCQ — Lead variants

**Step 3 calls lead variants with a frequency-dependent threshold: P < 5×10⁻⁸ for common variants
(MAF > 0.1%) but the stricter P < 6.25×10⁻¹⁰ for rare variants. The rare-variant threshold is
tightened because:**

- A. Rare variants have larger effect sizes, so they need a higher bar.
- B. There are far more rare-variant tests (~80M), so the standard 5×10⁻⁸ line is anti-conservative and would admit too many false positives.
- C. Rare variants are always artifacts and should be suppressed.
- D. LDSC requires it for the heritability estimate.

*Answer: **B.** The genome-wide 5×10⁻⁸ threshold is calibrated to ~1M independent common-variant
tests. Adding ~80M rare-variant tests inflates the multiple-testing burden, so a Bonferroni-style
tightening (0.05/8×10⁷) is needed to hold the family-wise error rate.*

**A follow-on on locus definition: this replication defines independent loci by distance (top SNP,
then a ±1 Mb exclusion window, iterated) rather than by the paper's LD-based r²≥0.05 merge. The main
scientific limitation of the distance-only rule is:**

- A. It cannot handle more than one chromosome.
- B. A single very strong signal (−log₁₀P ≫ 100) can generate association that extends beyond 2 Mb through LD, so distance clumping may split one true signal into several "independent" loci.
- C. It systematically undercounts loci relative to LD clumping.
- D. It requires an in-sample LD matrix to run.

*Answer: **B.** LD can carry a strong association past a fixed window; the paper's r²≥0.05 merge
exists specifically to stop that "bleeding" and collapse the fragments into one locus. Distance
clumping has no way to know two windows tag the same causal variant, so it can over-count. It is the
honest choice only when no matched genotype panel is available to compute pairwise r² — which is the
case here.*

# Fine-mapping + coding annotation (Step 4)

*SuSiE fine-mapping per locus (MHC excluded), then missense/splice annotation of credible-set
variants to nominate a candidate effector gene. **Caveat:** out-of-sample LD may miscalibrate
credible sets — interpret set sizes and PIPs with that in mind.*

```{r finemap-table}
annot_tab <- read_many(sm@input$annot)
show_table(annot_tab, "Credible-set variants and coding annotation")
```

Only the credible-set variants annotated as **missense** are shown below — these are the coding
candidates that fine-mapping nominates as potentially causal. The recovery of textbook LDL coding
variants (PCSK9 R46L, APOE C130R) at PIP = 1 is the strongest validation that the pipeline works
despite the out-of-sample LD.

```{r finemap-missense}
# Highlight the coding candidates. annot.tsv carries every credible-set variant; the missense
# subset is the biologically interpretable headline.
if (!is.null(annot_tab) && nrow(annot_tab) && "is_missense" %in% names(annot_tab)) {
  keep <- intersect(c("gene_symbol","ID","aa_change","pip","cs","sift_pred",
                      "polyphen_pred","is_known_causal"), names(annot_tab))
  miss <- annot_tab[is_missense == TRUE, ..keep][order(-pip)]
  show_table(miss, "Missense credible-set variants (coding effector candidates)")
} else {
  show_table(NULL)
}
```

```{r finemap-fig, results='asis'}
# Two cross-locus summary figures (per-locus QC counts; coding-vs-regulatory consequence mix).
show_figs(c(sm@input$finemap_summary, sm@input$consequence_summary), cols = 1)
```

```{r finemap-regional-fig, results='asis'}
# LocusZoom-style regional plots (GWAS -log10P coloured by r2-to-lead over a PIP track).
# We FEATURE three loci that each make a distinct point rather than showing all 11 — the
# two summary figures above already give the all-loci view, and 11 near-identical panels
# add scroll without insight:
#   PCSK9   - the R46L drug-target missense variant recovered at high PIP (headline result)
#   ABCG5_8 - two fine-mapped missense variants across a two-gene locus (coding + naming nuance)
#   APOE    - the canonical lipid locus with multiple credible sets (allelic heterogeneity)
featured <- c("PCSK9", "ABCG5_8", "APOE")
plotdir  <- sm@input$finemap_plots[1]
regional_pngs <- file.path(plotdir, sprintf("%s.%s.regional.png", "LDL_C", featured))
cat("*Three representative loci (of 11 fine-mapped). PCSK9 shows the R46L drug-target coding",
    "variant at high PIP; ABCG5/ABCG8 shows two missense variants across a two-gene locus;",
    "APOE shows a locus resolved into multiple credible sets. The remaining loci are summarised",
    "in the two figures above and available in the repository.*\n\n")
show_figs(regional_pngs, cols = 1)
```

## MCQ — Fine-mapping

**The pipeline fine-maps with an out-of-sample 1000G EUR LD reference rather than the in-sample
UKBB LD used by the reference paper. The primary risk this introduces to the credible sets is:**

- A. They will always be too small.
- B. LD mismatch can miscalibrate posterior inclusion probabilities and yield spurious or oversized credible sets.
- C. SuSiE cannot run at all without in-sample LD.
- D. It only affects the Manhattan plot, not fine-mapping.

*Answer: **B.** SuSiE-RSS assumes the LD matrix matches the GWAS sample; a mismatched reference
distorts PIPs and credible-set membership. In this run the fingerprint is visible in the
convergence QC (most loci hit the IBSS iteration cap), which is why an `estimate_s_rss` LD-mismatch
diagnostic is reported per locus and `estimate_residual_variance` is fixed to `FALSE` — the
documented remedy for out-of-sample LD. This is the single most important caveat of the analysis.*

# Limitations

- **Out-of-sample LD.** Fine-mapping and distance-clumping use a 1000G EUR reference rather than
  in-sample LD; SuSiE credible sets may be miscalibrated (flagged per locus by the kriging
  diagnostic). This is the single most important caveat and the main methodological divergence
  from the reference paper, which had in-sample UKBB LD.
- **Distance-only locus definition.** With no genotype panel provided at Step 3, independent loci
  are defined by a ±1 Mb distance rule rather than the paper's r²≥0.05 LD merge; a very strong
  signal can bleed past the window and be split into apparent multiple loci (see the Step-3 MCQ).
- **Fine-mapping scoped to LDL effector loci.** Step 4 fine-maps the 11 canonical LDL effector
  loci (a hypothesis-driven subset), not genome-wide — chosen so out-of-sample-LD behaviour can be
  inspected locus by locus.
- **Coding annotation explains few loci.** Of the credible-set variants, only a handful are
  missense; most signal is regulatory/intronic. Resolving those mechanisms needs molecular-QTL
  colocalization — the next step, not yet included here.
- **EUR-only.** Exposures and (planned) outcomes are European-ancestry, so estimates may not
  generalize — a deliberate choice to keep LD and allele frequencies matched.

*Steps 5–7 (colocalization, Mendelian randomization) are in progress and will be added to this
report, each with its own MCQ, once their pipeline rules have run.*

