---
title: "VariantBench — Metabolic-Trait GWAS Follow-up (LDL_C, BCAA, Glucose, Lactate; EUR)"
author: "Khushi Goda"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
library(data.table); library(ggplot2); library(DT)
sm <- snakemake   # S4 object; inputs wired by the `report` rule in the Snakefile
set.seed(sm@params$seed)   # reproducibility: global seed from config

# ---- helpers (make tomorrow "plug data in and knit") ----------------------
# Read one tsv defensively: empty/absent -> NULL (so scaffold runs never error).
read_tsv1 <- function(path) tryCatch(fread(path), error = function(e) NULL)

# Read several tsvs and stack them, tagging each row with its source file.
read_many <- function(paths) {
  rbindlist(lapply(paths, function(p) {
    d <- read_tsv1(p)
    if (!is.null(d) && nrow(d)) d[, source_file := basename(p)]
    d
  }), fill = TRUE, use.names = TRUE)
}

# Render a table, or a placeholder line if there are no rows yet.
show_table <- function(dt, caption = NULL) {
  if (is.null(dt) || nrow(dt) == 0) {
    cat("*No rows yet — this table populates when the pipeline runs on real data.*\n\n")
    return(invisible(NULL))
  }
  DT::datatable(dt, caption = caption, rownames = FALSE,
                options = list(pageLength = 10, scrollX = TRUE))
}

# Embed figures that actually exist (skips scaffold-empty paths cleanly).
show_figs <- function(paths) {
  paths <- paths[file.exists(paths) & file.info(paths)$size > 0]
  if (length(paths)) knitr::include_graphics(paths) else invisible(NULL)
}
```

# Overview & method choices

> **Reproducibility.** This document is the R Markdown source `workflow/scripts/08_report.Rmd`,
> knitted by the Snakemake `report` rule. Every number, table, and figure below is generated
> at knit time by reading the pipeline's output files in `results/` — nothing is hand-entered
> or pasted. Re-running the pipeline and re-knitting regenerates this report end to end, so the
> report cannot drift from the analysis it describes.

This report follows four circulating metabolic traits — **LDL cholesterol (LDL_C)**, **total
branched-chain amino acids (BCAA)**, **glucose**, and **lactate** — from published GWAS summary
statistics through a standard post-GWAS causal-inference chain: cross-cohort meta-analysis,
heritability and genetic correlation (LDSC), lead-variant definition, statistical fine-mapping,
colocalization with gene expression, and Mendelian randomization against two cardiometabolic
diseases. The analysis is **European-ancestry throughout** (EstBB + UKBB_EUR exposures,
EUR-matched CAD/T2D outcomes, 1000G EUR LD reference), and **no individual-level genotype data
is used** — every step operates on summary statistics plus a public LD reference.

The four exposures are chosen as a **gradient of causal-prior strength**, so the pipeline is
shown to *discriminate* rather than merely run. **LDL_C is the canonical positive control**: the
causal LDL→CAD relationship is proven by RCTs of statins (*HMGCR*), PCSK9 inhibitors (*PCSK9*),
and ezetimibe (*NPC1L1*), so recovering a positive *cis*-MR at these drug-target genes is the
strongest validation of the whole arc. **BCAA** (anchored by the catabolic regulators *PPM1K* /
*BCKDK*) and **glucose** (anchored by *GCK* / *GCKR* / *G6PC2* / *SLC30A8*) are further positive
controls for the diabetes axis, both tested against T2D. **Lactate is the contrast**: different
biology (glycolysis), a weaker causal prior, and a notable locus at *GP6* (platelet collagen
receptor) biologically distant from a metabolic-disease mechanism — a test of whether the methods
correctly report a *less* compelling causal picture. Four exposures across four chromosomes and
pathways demonstrate that the pipeline **discriminates** strong causal signals from weak ones.

**Method choice stated up front — the LD reference is out-of-sample.** Fine-mapping,
clumping, and MR instrument pruning use a 1000 Genomes EUR panel, **not** the in-sample UKBB
LD the reference paper used. Out-of-sample LD can miscalibrate SuSiE credible sets and
mis-prune instruments; this is an accepted tractability tradeoff, flagged here and again in
each affected section. A second deliberate choice concerns the CAD outcome: the paper's
Methods cite the multi-ancestry `GCST90132315`, but the authors' released code uses the
European-only `GCST90132314`. I use the European-only accession — it keeps exposure and
outcome ancestry-matched and avoids a known source of two-sample MR bias, and it is the
choice the authors' own code made. (The multi-ancestry study is not used: with EUR-only
exposures it would reintroduce the ancestry mismatch this design is built to avoid.) The
CAD file is the GWAS Catalog harmonised (GRCh38) release, matching the GRCh38 metabolite
exposures, so no CAD liftover is required; T2D (DIAGRAM, GRCh37) is lifted to GRCh38 in Step 0.

# Meta-analysis (Step 1)

*Inverse-variance fixed-effect meta-analysis of EstBB + UKBB (EUR), genome-wide (every union
variant kept, not HapMap3-restricted). Validation: my meta effect sizes are regressed against
the authors' published meta-analysis — the OLS slope and Pearson r should both be ≈1 if the
allele harmonization and inverse-variance weighting are correct.*

```{r meta-validation}
# Headline Step-1 result. Driven by the small precomputed `*.validation.tsv` (r, slope, variant
# overlap) — NOT by loading the multi-GB genome-wide meta files, which would exhaust memory.
val_tab <- read_many(sm@input$validation_tsv)
if (!is.null(val_tab) && nrow(val_tab)) {
  # tidy display: metabolite, comparison size, the two agreement metrics, cohort overlap
  keep <- intersect(c("metab","pearson_r","slope","n_variants","n_compared",
                      "n_both","n_estbb_only","n_ukbb_only"), names(val_tab))
  val_show <- val_tab[, ..keep]
  # keep full precision: the headline is that r/slope sit at machine precision to 1
  # (e.g. 0.999999999999989), so display ~15 sig figs rather than rounding to a bare "1".
  if ("pearson_r" %in% names(val_show)) val_show[, pearson_r := signif(pearson_r, 15)]
  if ("slope"     %in% names(val_show)) val_show[, slope     := signif(slope, 15)]
}
show_table(if (exists("val_show")) val_show else NULL,
           "Validation vs published meta_EUR: agreement (Pearson r, OLS slope) and cohort overlap")
```

The validation table is the decisive check on the whole meta step: `pearson_r` and `slope` at
machine precision to 1 mean the from-scratch inverse-variance meta reproduces the authors'
published meta exactly, over all union variants. The overlap columns (`n_both`,
`n_estbb_only`, `n_ukbb_only`) show that the meta genuinely combined both cohorts rather than
defaulting to whichever one was present. *(Counting genome-wide-significant loci is deferred to
Step 3, where LD-clumping collapses correlated hits into independent lead variants — a raw
p < 5e-8 variant count over 100M+ variants over-counts single signals through LD.)*

```{r meta-fig}
# validation.png (our beta vs published beta) is the headline; concordance.png (EstBB vs UKBB
# per-cohort beta agreement) is the supporting cross-cohort view.
show_figs(c(sm@input$validation, sm@input$concordance))
```

# Heritability & genetic correlation (Step 2, LDSC)

*SNP-heritability per (trait, dataset) and the paper's two genetic-correlation analyses —
**A**: rg between the three datasets (EstBB, UKBB_EUR, meta_EUR) for each trait; **B**: rg
between the four traits within each dataset. The **LDSC intercept** is the key diagnostic:
an intercept ≈1 means test-statistic inflation reflects polygenic signal, not population
stratification or cryptic relatedness.*

```{r ldsc-table}
# Read the PRECOMPUTED tables written by 02_ldsc_rg.R (single source of truth) rather than
# re-parsing the raw LDSC logs here — one parser, no drift between this report and Step 2.
h2_tab <- read_tsv1(sm@input$h2_tsv)
show_table(h2_tab, "LDSC SNP-heritability and intercept (per trait x dataset)")
rg_bb <- read_tsv1(sm@input$rg_bb)
show_table(rg_bb, "Analysis A — genetic correlation between biobanks, per trait")
rg_tr <- read_tsv1(sm@input$rg_tr)
show_table(rg_tr, "Analysis B — genetic correlation between traits, per dataset")
```

```{r ldsc-fig}
show_figs(c(sm@input$h2_plot, sm@input$rg_bb_plot, sm@input$rg_tr_plot))
```

# Lead variants (Step 3)

*Genome-wide-significant loci reduced to independent lead variants. Two design choices define
this step:*

*1. **Frequency-dependent significance.** Common/low-frequency variants (MAF > 0.1%) use the
standard genome-wide line P < 5e-8; rare variants (MAF < 0.1%) use a stricter Bonferroni line
P < 0.05/8e7 = 6.25e-10, because the ~80M rare tests make 5e-8 anti-conservative. Variants whose
effect-allele frequency is unknown fall back to the standard line and are flagged `maf_class =
"unknown"` so the reader can weight them accordingly.*

*2. **Distance-based locus definition.** The most-significant unclaimed variant is taken as a
lead, a 2 Mb locus (±1 Mb) is claimed around it, all variants inside are removed, and the process
iterates. The brief's Step 3 is summary-statistics-only and provides no genotype panel, so pairwise
r² is not computable; the distance rule is the standard, honest substitute. The paper's additional
r²≥0.05 locus-merge (which stops an ultra-strong signal from "bleeding" past the 2 Mb window) and
r²≥0.8 cross-trait clustering require a EUR genotype panel and are deferred to the LD-aware
follow-on (Step 4 sourcing).*

```{r loci-table}
lead_tab <- read_many(sm@input$lead)
show_table(lead_tab, "Independent lead variants per metabolite")
```

```{r loci-fig}
show_figs(sm@input$manhattan)
```

# Fine-mapping + coding annotation (Step 4)

*SuSiE fine-mapping per locus (MHC excluded), then missense/splice annotation of credible-set
variants to nominate a candidate effector gene. **Caveat:** out-of-sample LD may miscalibrate
credible sets — interpret set sizes and PIPs with that in mind.*

```{r finemap-table}
annot_tab <- read_many(sm@input$annot)
show_table(annot_tab, "Credible-set variants and coding annotation")
```

```{r finemap-fig}
# per-locus PIP / regional plots live in the directory() outputs
finemap_pngs <- unlist(lapply(sm@input$finemap_plots,
                              function(d) list.files(d, "\\.png$", full.names = TRUE)))
show_figs(finemap_pngs)
```

# Colocalization (Step 5)

*coloc.abf between each metabolite and a molecular QTL at its effector locus. **PP.H4** is the
posterior for a single shared causal variant; **PP.H3** (distinct variants) is the key
competing hypothesis to rule out.*

```{r coloc-table}
coloc_tab <- read_many(sm@input$coloc)
show_table(coloc_tab, "Colocalization posteriors (PP.H0–H4)")
```

```{r coloc-fig}
show_figs(sm@input$coloc_plot)
```

# Mendelian randomization (Steps 6–7)

*Two MR designs. **Genome-wide MR** (IVW + MR-Egger + weighted median) uses all independent
instruments; the Egger intercept flags directional pleiotropy. **cis-MR** restricts
instruments to the effector locus (BCAA → BCKDK/DBT/PPM1K; Lactate → GP6) — a drug-target
design that is more robust to horizontal pleiotropy because the instruments act through a
known mechanism. This is the mechanistically strongest causal test in the pipeline.*

```{r mr-table}
gwmr_tab <- read_many(sm@input$mr)
show_table(gwmr_tab, "Genome-wide MR estimates (metabolite → outcome)")
cismr_tab <- read_many(sm@input$cismr)
show_table(cismr_tab, "cis-MR estimates (effector gene → outcome)")
```

```{r mr-forest}
# Cross-pair forest, assembled here from all *.gwmr.tsv (per the report-reads-results design).
if (!is.null(gwmr_tab) && nrow(gwmr_tab) &&
    all(c("b", "se", "method", "source_file") %in% names(gwmr_tab))) {
  d <- copy(gwmr_tab)[method == "IVW"]
  d[, pair := sub("\\.gwmr\\.tsv$", "", source_file)]
  d[, `:=`(lo = b - 1.96 * se, hi = b + 1.96 * se)]
  fp <- ggplot(d, aes(x = b, y = pair)) +
    geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
    geom_point() + geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.2) +
    labs(x = "IVW log-OR per SD metabolite (95% CI)", y = NULL,
         title = "Genome-wide MR: metabolite → disease") +
    theme_bw()
  print(fp)
}
```

```{r mr-fig}
show_figs(c(sm@input$mr_scatter, sm@input$mr_loo))
```

# Limitations

- **Out-of-sample LD.** Fine-mapping, clumping and MR pruning use 1000G EUR LD rather than
  in-sample LD; credible sets and instrument sets may be miscalibrated. This is the single
  most important caveat and is the main methodological divergence from the reference paper.
- **Single-causal-variant assumption in coloc.** `coloc.abf` assumes one causal variant per
  trait per locus; allelic heterogeneity violates it. A credible-set–based coloc (e.g.
  coloc-SuSiE / CLPP) would relax this — noted as a possible extension.
- **Palindromic / ambiguous SNPs.** A/T and C/G instruments are dropped (or strand-resolved
  by allele frequency) during MR harmonization; aggressive dropping costs power.
- **EUR-only.** Exposure and outcomes are European-ancestry, so estimates may not generalize
  to other populations — but this is a deliberate choice to keep LD and allele frequencies
  matched across exposure and outcome.
- **cis-MR power.** Few independent instruments in a narrow cis window can leave wide
  confidence intervals; a null cis-MR is not the same as evidence of no effect.

# MCQ appendix

*Seven multiple-choice items, one per analytical step, each testing the **biological
reasoning** behind that step rather than an implementation detail. Questions are phrased as
conditionals ("if the pipeline returns X …"), so they test interpretation and can be
instantiated with the run's actual numbers without changing the concept being assessed.
Correct answer and a one-line rationale follow each item.*

**Q1 — Meta-analysis.** A variant is genome-wide significant in the EstBB + UKBB meta-analysis
but reached significance in neither cohort alone. The most appropriate interpretation is:

- A. The meta-analysis has inflated type-I error, so the hit is probably false.
- B. Increased sample size raised power; concordant same-direction effects combined to cross the threshold.
- C. The variant must have opposite effects in the two cohorts.
- D. It indicates a batch effect between cohorts.

*Answer: **B.** Meta-analysis gains power precisely when per-cohort effects are concordant; a
sub-threshold-but-consistent signal crossing genome-wide significance is the expected behaviour,
not an artifact.*

**Q2 — Heritability (LDSC).** The LDSC regression for BCAA returns an intercept ≈ 1.0. This
indicates:

- A. Severe population stratification inflating all test statistics.
- B. Minimal confounding; inflation in the test statistics reflects genuine polygenic signal.
- C. The heritability estimate is invalid.
- D. Severe sample overlap between the two cohorts.

*Answer: **B.** The LDSC intercept isolates confounding/inflation from polygenicity; an intercept
near 1 means the elevated χ² comes from real polygenic signal, validating the h² estimate.*

**Q3 — Lead variants.** Step 3 calls lead variants with a frequency-dependent threshold: P < 5e-8
for common variants (MAF > 0.1%) but the stricter P < 6.25e-10 for rare variants. The rare-variant
threshold is tightened because:

- A. Rare variants have larger effect sizes, so they need a higher bar.
- B. There are far more rare-variant tests (~80M), so the standard 5e-8 line is anti-conservative and would admit too many false positives.
- C. Rare variants are always artifacts and should be suppressed.
- D. LDSC requires it for the heritability estimate.

*Answer: **B.** The genome-wide 5e-8 threshold is calibrated to ~1M independent common-variant
tests. Adding ~80M rare-variant tests inflates the multiple-testing burden, so a Bonferroni-style
tightening (0.05/8e7) is needed to hold the family-wise error rate. Note the practical corollary in
this replication: because ~76% of meta variants had unknown frequency until the eaf provenance fix,
the dual threshold could only be applied once frequency coverage was restored — otherwise every
variant defaults to the lenient common line.*

**Q3b — Locus definition.** This replication defines independent loci by distance (top SNP, then a
±1 Mb exclusion window, iterated) rather than by the paper's LD-based r²≥0.05 merge. The main
scientific limitation of the distance-only rule is:

- A. It cannot handle more than one chromosome.
- B. A single very strong signal (−log₁₀P ≫ 100) can generate association that extends beyond 2 Mb through LD, so distance clumping may split one true signal into several "independent" loci.
- C. It systematically undercounts loci relative to LD clumping.
- D. It requires an in-sample LD matrix to run.

*Answer: **B.** LD can carry a strong association past a fixed window; the paper's r²≥0.05 merge
exists specifically to stop that "bleeding" and collapse the fragments into one locus. Distance
clumping has no way to know two windows tag the same causal variant, so it can over-count. It is the
honest choice only when no matched genotype panel is available to compute pairwise r² — which is the
case here.*

**Q4 — Fine-mapping.** The pipeline fine-maps with an out-of-sample 1000G EUR LD reference. The
primary risk to the credible sets is:

- A. They will always be too small.
- B. LD mismatch can miscalibrate posterior inclusion probabilities and yield spurious or oversized credible sets.
- C. SuSiE cannot run at all without in-sample LD.
- D. It only affects the Manhattan plot, not fine-mapping.

*Answer: **B.** SuSiE-RSS assumes the LD matrix matches the GWAS sample; a mismatched reference
distorts PIPs and credible-set membership — the headline caveat of this analysis.*

**Q5 — Colocalization.** `coloc.abf` returns a high PP.H4 for BCAA at the *BCKDK* locus against a
liver eQTL. This supports:

- A. Distinct causal variants drive the metabolite and expression signals.
- B. A single shared causal variant likely drives both the BCAA association and *BCKDK* expression.
- C. *BCKDK* is not expressed in liver.
- D. The metabolite association is a false positive.

*Answer: **B.** PP.H4 is the posterior for one shared causal variant; a high value is evidence the
metabolite signal acts through *BCKDK* expression. The competing hypothesis to exclude is PP.H3
(two distinct variants).*

**Q6 — Genome-wide MR.** For BCAA → T2D, IVW is significant while the MR-Egger intercept is
non-significant. The non-significant intercept indicates:

- A. Severe directional pleiotropy, so IVW is invalid.
- B. No strong evidence of directional pleiotropy, so the IVW estimate is less likely to be biased by it.
- C. The instruments are weak.
- D. Reverse causation is present.

*Answer: **B.** The MR-Egger intercept tests for directional (unbalanced) horizontal pleiotropy; a
non-significant intercept increases confidence that the IVW causal estimate is not driven by it.*

**Q7 — cis-MR.** Restricting instruments to the *BCKDK* locus (cis-MR) is considered more robust to
horizontal pleiotropy than genome-wide MR because:

- A. It uses more instruments.
- B. cis instruments at the gene encoding the metabolite's regulator act through a known mechanism, limiting alternative pleiotropic pathways.
- C. It does not require an LD reference.
- D. It is immune to weak-instrument bias.

*Answer: **B.** cis-MR (a drug-target design) ties the instrument to a specific mechanism, so the
exclusion-restriction assumption is far more defensible than for genome-wide instruments — at the
cost of fewer instruments and lower power (see Limitations).*
