---
title: "From GWAS Summary Statistics to Candidate Causal Genes: A Reproducible Post-GWAS Pipeline for Circulating Metabolic Traits in European-Ancestry Biobanks"
subtitle: "Meta-analysis, heritability, fine-mapping, colocalization, and Mendelian randomization for LDL cholesterol, branched-chain amino acids, glucose, and lactate"
author: "Khushi Goda"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    df_print: paged
    number_sections: true
---

---

> **GitHub Repository:** https://github.com/khushigoda/variantbench  
> **Analysis Methods:** [ANALYSIS.md](../ANALYSIS.md)

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
library(data.table); library(ggplot2); library(DT)
library(cowplot); library(png)   # compose_figs(): tile per-trait PNGs into one image
sm <- snakemake   # S4 object; inputs wired by the `report` rule in the Snakefile
set.seed(sm@params$seed)   # reproducibility: global seed from config

# ---- helpers (make tomorrow "plug data in and knit") ----------------------
# Read one tsv defensively: empty/absent -> NULL (so scaffold runs never error).
read_tsv1 <- function(path) tryCatch(fread(path), error = function(e) NULL)

# Read several tsvs and stack them, tagging each row with its source file.
read_many <- function(paths) {
  rbindlist(lapply(paths, function(p) {
    d <- read_tsv1(p)
    if (!is.null(d) && nrow(d)) d[, source_file := basename(p)]
    d
  }), fill = TRUE, use.names = TRUE)
}

# Render a table, or a placeholder line if there are no rows yet.
show_table <- function(dt, caption = NULL) {
  if (is.null(dt) || nrow(dt) == 0) {
    cat("*No rows yet — this table populates when the pipeline runs on real data.*\n\n")
    return(invisible(NULL))
  }
  DT::datatable(dt, caption = caption, rownames = FALSE,
                options = list(pageLength = 10, scrollX = TRUE))
}

# Embed figures that actually exist (skips scaffold-empty paths cleanly).
# NB: we DON'T use include_graphics here. Snakemake renders the Rmd from a temp
# script dir (.snakemake/scripts/), and both include_graphics and Pandoc's
# self_contained pass resolve image paths relative to THAT dir — so any path we pass
# (relative -> "Could not fetch resource"; absolute -> relativised to "../../results/..")
# fails. Instead we read each file (cwd = knit_root_dir = repo root, so file.exists
# passes) and inline it as a base64 data-URI <img>, which needs no path resolution at
# all. Chunks calling this MUST set results='asis' so the HTML is emitted raw.
show_figs <- function(paths, cols = NULL) {
  paths <- paths[file.exists(paths) & file.info(paths)$size > 0]
  if (!length(paths)) return(invisible(NULL))
  # A single figure renders full-width & centred; multiple figures render in a
  # responsive CSS grid (default 2 cols, auto-fit down to 1 on narrow screens) so
  # groups like the 11 regional LocusZoom panels read as a compact gallery instead
  # of an endless vertical scroll.
  if (is.null(cols)) cols <- if (length(paths) > 1) 2 else 1
  if (cols <= 1 || length(paths) == 1) {
    for (p in paths)
      cat(sprintf('<img src="%s" style="max-width:100%%;display:block;margin:0.5em auto;" alt="%s"/>\n\n',
                  knitr::image_uri(p), basename(p)))
  } else {
    cat(sprintf('<div style="display:grid;grid-template-columns:repeat(%d,1fr);gap:0.6em;margin:0.5em 0;">\n', cols))
    for (p in paths)
      cat(sprintf('<img src="%s" style="width:100%%;height:auto;" alt="%s"/>\n',
                  knitr::image_uri(p), basename(p)))
    cat('</div>\n\n')
  }
  invisible(NULL)
}

# Tile several PNGs into ONE composite image (cowplot) and inline it base64.
# Used for the Step-1 per-trait scatter families (4 validation, 4 concordance):
# one 2x2 image reads far better than 8 separate full-width plots and cuts weight.
# Each source PNG already carries its own trait title, so no cowplot labels.
compose_figs <- function(paths, ncol = 2, panel_in = 4.5) {
  paths <- paths[file.exists(paths) & file.info(paths)$size > 0]
  if (!length(paths)) return(invisible(NULL))
  if (length(paths) == 1) return(show_figs(paths))
  grobs <- lapply(paths, function(p)
    ggdraw() + draw_grob(grid::rasterGrob(readPNG(p), interpolate = TRUE)))
  g <- plot_grid(plotlist = grobs, ncol = ncol)
  nrow <- ceiling(length(paths) / ncol)
  tmp <- tempfile(fileext = ".png")
  ggsave(tmp, g, width = panel_in * ncol, height = panel_in * nrow,
         dpi = 150, device = grDevices::png, type = "cairo", bg = "white")
  cat(sprintf('<img src="%s" style="max-width:100%%;display:block;margin:0.5em auto;" alt="composite"/>\n\n',
              knitr::image_uri(tmp)))
  invisible(NULL)
}
```

# Overview & method choices

> **Reproducibility.** This document is the R Markdown source `workflow/scripts/08_report.Rmd`,
> knitted by the Snakemake `report` rule. Every number, table, and figure below is generated
> at knit time by reading the pipeline's output files in `results/` — nothing is hand-entered
> or pasted. Re-running the pipeline and re-knitting regenerates this report end to end, so the
> report cannot drift from the analysis it describes.

> **Scope of this knit.** This version reports the **completed steps: meta-analysis, LDSC,
> lead-variant definition, and fine-mapping + coding annotation (Steps 1–4)**, each followed by a
> multiple-choice question testing the biological reasoning behind that step. Steps 1–2 run across
> all four traits; Steps 3–4 are shown for **LDL_C**, the positive-control trait whose 11 canonical
> effector loci let the out-of-sample-LD behaviour be inspected locus by locus. Colocalization with
> molecular QTLs (Step 5) and Mendelian randomization — genome-wide and drug-target *cis*-MR
> (Steps 6–7) — are also shown for **LDL_C**, each step with its own MCQ, testing the causal effect
> of LDL-C on coronary artery disease.

This report follows four circulating metabolic traits — **LDL cholesterol (LDL_C)**, **total
branched-chain amino acids (BCAA)**, **glucose**, and **lactate** — from published GWAS summary
statistics through a standard post-GWAS causal-inference chain: cross-cohort meta-analysis,
heritability and genetic correlation (LDSC), lead-variant definition, statistical fine-mapping,
colocalization with gene expression, and Mendelian randomization against two cardiometabolic
diseases. The analysis is **European-ancestry throughout** (EstBB + UKBB_EUR exposures,
EUR-matched CAD/T2D outcomes, 1000G EUR LD reference), and **no individual-level genotype data
is used** — every step operates on summary statistics plus a public LD reference.

**Data provenance.** All exposure summary statistics come from Tambets *et al.*[^tambets], a GWAS
meta-analysis of 249 circulating metabolic traits in the **Estonian Biobank (EstBB)** and the
**UK Biobank (UKBB)**, retrieved as GWAS-SSF-formatted files from the **NHGRI-EBI GWAS Catalog**.
For each metabolite the pipeline uses the two per-cohort European releases (EstBB and UKBB_EUR) as
meta-analysis inputs and the authors' own published European meta-analysis as the Step-1 validation
target. The **coronary artery disease (CAD)** outcome used downstream is the European-ancestry
release of Aragam *et al.*[^aragam] (GWAS Catalog `GCST90132314`). **Every file is on GRCh38**: the
metabolite and CAD statistics are served already harmonised to GRCh38 by the GWAS Catalog, so no
coordinate liftover is applied to them.

The four exposures are chosen as a **gradient of causal-prior strength**, so the pipeline is
shown to *discriminate* rather than merely run. **LDL_C is the canonical positive control**: the
causal LDL→CAD relationship is established by the totality of genetic, epidemiologic, and
randomized-trial evidence — including RCTs of statins (*HMGCR*), PCSK9 inhibitors (*PCSK9*), and
ezetimibe (*NPC1L1*)[^ference] — so recovering a positive *cis*-MR at these drug-target genes is
the strongest validation of the whole arc. **BCAA** (anchored by the catabolic regulators *PPM1K* /
*BCKDK*), whose causal role in type 2 diabetes has itself been supported by Mendelian
randomization[^lotta], and **glucose** (anchored by *GCK* / *GCKR* / *G6PC2* / *SLC30A8*) are
further positive controls for the diabetes axis, both tested against T2D. **Lactate is the
contrast**: different biology (glycolysis), a weaker causal prior, and a notable locus at *GP6*
(platelet collagen receptor) biologically distant from a metabolic-disease mechanism — a test of
whether the methods correctly report a *less* compelling causal picture. Together the four exposures
span loci across the genome and several distinct pathways, so the pipeline can be judged on whether
it **discriminates** strong causal signals from weak ones.

[^tambets]: Tambets R, *et al.* Genome-wide association study for circulating metabolic traits in
619,372 individuals. *medRxiv* 2024.10.15.24315557 (2024); published in *Nature* (2026),
doi:10.1038/s41586-026-10532-5. Summary statistics via the NHGRI-EBI GWAS Catalog.

[^aragam]: Aragam KG, *et al.* Discovery and systematic characterization of risk variants and genes
for coronary artery disease in over a million participants. *Nat Genet* 54, 1803–1815 (2022),
doi:10.1038/s41588-022-01233-6. GWAS Catalog `GCST90132314` (European-ancestry release).

[^ference]: Ference BA, *et al.* Low-density lipoproteins cause atherosclerotic cardiovascular
disease. 1. Evidence from genetic, epidemiologic, and clinical studies. A consensus statement from
the European Atherosclerosis Society Consensus Panel. *Eur Heart J* 38(32), 2459–2472 (2017),
doi:10.1093/eurheartj/ehx144.

[^lotta]: Lotta LA, *et al.* Genetic predisposition to an impaired metabolism of the branched-chain
amino acids and risk of type 2 diabetes: a Mendelian randomisation analysis. *PLoS Med* 13(11),
e1002179 (2016), doi:10.1371/journal.pmed.1002179.

**Scope and resource constraints, stated up front.** Because of finite compute and a fixed
time budget, the analysis was **not** run across the full planned set of metabolites — the
report focuses on LDL_C (with the other three traits carried through the early steps) rather than
the paper's full metabolite panel. This is a scoping decision, not an architectural limit: the
pipeline is **config-driven** (`config/config.yaml`), so adding a metabolite is a one-line
accession entry and the same Snakemake rules then run end to end for it — the design scales to any
desired number of traits, and the restriction here is purely one of available resources. For the
same reason, some **tools and methodological choices deliberately diverge from the published
study** where a lighter or more self-contained alternative fit the resource envelope (these
substitutions are flagged in the affected sections). All work runs on a single workstation-class
machine (**8-core CPU, 8 GiB RAM, no GPU**); memory in particular drives several of these choices.

**The LD reference is out-of-sample.** Fine-mapping and MR instrument pruning use a
1000 Genomes EUR panel, **not** the in-sample UKBB LD the reference paper used. Out-of-sample LD
can miscalibrate SuSiE credible sets and mis-prune instruments; this is an accepted tractability
tradeoff, flagged here and again in each affected section. With LDL_C selected as the metabolite of interest, **CAD is the
appropriate disease outcome** for the downstream colocalization and MR, so T2D is left out of the
reported arc.

# Meta-analysis (Step 1)

This step combines the two European cohorts — EstBB and UKBB_EUR — into a single genome-wide
association table for each metabolite using **inverse-variance-weighted, fixed-effect
meta-analysis**. The main steps are: each cohort's summary statistics are matched variant by
variant on chromosome, position, and both alleles (a **no-flip join**, so only identically
oriented effect alleles are combined and any strand or allele mismatch is left uncombined rather
than silently mis-merged); every variant is weighted by the inverse of its squared standard error,
so more precisely estimated cohorts contribute more; the pooled effect is the weighted mean of the
per-cohort betas, with a pooled standard error of √(1/Σw); and effect-allele frequency and sample
size are carried through, with a signed *z*-statistic, two-sided *p*-value, and a Cochran's-*Q*
heterogeneity test computed per variant. Crucially, **every union variant is kept** — the table is
not restricted to a HapMap3 or well-imputed subset — because the downstream steps (lead-variant
definition, fine-mapping, MR) need the full genome-wide signal. The whole meta is run
**one chromosome at a time and written incrementally**, so the full ~100M-variant table is produced
without ever holding more than a single chromosome in memory — the design choice that lets a
genome-wide meta run on an 8 GiB machine. Finally, the result is **validated against the authors'
published `meta_EUR`**: our pooled betas are regressed on theirs, and the OLS slope and Pearson *r*
should both sit at ≈1 if the allele harmonization and inverse-variance weighting are correct.

```{r meta-validation}
# Headline Step-1 result. Driven by the small precomputed `*.validation.tsv` (r, slope, variant
# overlap) — NOT by loading the multi-GB genome-wide meta files, which would exhaust memory.
val_tab <- read_many(sm@input$validation_tsv)
if (!is.null(val_tab) && nrow(val_tab)) {
  # tidy display: metabolite, comparison size, the two agreement metrics, cohort overlap
  keep <- intersect(c("metab","pearson_r","slope","n_variants","n_compared",
                      "n_both","n_estbb_only","n_ukbb_only"), names(val_tab))
  val_show <- val_tab[, ..keep]
  # keep full precision: the headline is that r/slope sit at machine precision to 1
  # (e.g. 0.999999999999989), so display ~15 sig figs rather than rounding to a bare "1".
  if ("pearson_r" %in% names(val_show)) val_show[, pearson_r := signif(pearson_r, 15)]
  if ("slope"     %in% names(val_show)) val_show[, slope     := signif(slope, 15)]
}
show_table(if (exists("val_show")) val_show else NULL,
           "Validation vs published meta_EUR: agreement (Pearson r, OLS slope) and cohort overlap")
```

The table reads out that validation. `pearson_r` and `slope` sit at machine precision to 1, meaning
the pooled effect sizes reproduce the authors' published meta exactly across all union variants,
while the overlap columns (`n_both`, `n_estbb_only`, `n_ukbb_only`) confirm the meta genuinely
combined both cohorts rather than defaulting to whichever one was present. Together these confirm
that the summary-statistic pre-processing was accurate and the genome-wide table is ready for the
downstream steps.

**How this compares to the source manuscript.** The reproduced European meta-analysis pools the
same two cohorts the authors describe — EstBB (n ≈ 185,352) and UKBB_EUR (n ≈ 413,897) — into a
`meta_EUR` of **n = 599,249**, matching the published sample size exactly. The validation scatter
here plays the same role as the manuscript's Supplementary Fig. 2, which regresses shared
lead-variant betas from their `meta_EUR` against a prior published analysis (Karjalainen et al.)
and reports near-perfect concordance where the samples overlap. The comparison differs in one
honest respect: the manuscript benchmarks against an *external* study with only partial sample
overlap, so their agreement is high but sub-unity; this replication benchmarks against the authors'
*own* published `meta_EUR` over the identical cohorts, so `r` and slope sit at machine precision to
1 — a reproduction check rather than a cross-study replication. Both confirm the same thing: allele
harmonisation and inverse-variance weighting recover effect sizes that agree with the reference.

```{r meta-fig, results='asis'}
# Two 2x2 composites (one image each) instead of 8 stacked plots.
# Validation (meta_rep vs meta_og) is the headline reproducibility check;
# concordance (EstBB vs UKBB per-cohort beta) is the supporting cross-cohort view.
cat("**Validation — reproduced meta (`meta_rep`) against the original published meta (`meta_og`), per metabolite:**\n\n")
compose_figs(sm@input$validation)
cat("**Cross-cohort concordance — EstBB vs UKBB effect sizes, per metabolite:**\n\n")
compose_figs(sm@input$concordance)
```

**Limitations.** This is a fixed-effect, inverse-variance-weighted meta-analysis, which assumes the
two cohorts estimate the same underlying effect. The per-variant Cochran's-*Q* is carried through so
heterogeneous loci can be flagged, but a fixed-effect pooled estimate can still be misleading where
the true effect genuinely differs between EstBB and UKBB. The no-flip join is deliberately
conservative — strand-ambiguous or allele-mismatched variants are left uncombined rather than risk a
silent mis-merge — so a small number of genuinely shared variants that a more aggressive
allele-flipping step could have harmonised are instead reported from a single cohort. Finally, the
validation here is a *reproduction* check against the authors' own published meta over the identical
cohorts: near-perfect agreement confirms the allele harmonisation and inverse-variance weighting are
correct, but it does not independently replicate the association signal in a fresh sample the way a
comparison against a separate study would. Nor does genome-wide reproduction validate every locus
individually: a pooled hit with opposite-signed cohort effects and strong Cochran's-*Q* heterogeneity
should still be treated as potentially cohort-specific or artefactual rather than automatically as a
shared biological association.

## MCQ — Meta-analysis

**Two cohorts can be pooled for very different reasons: one cohort may simply dominate the result, or both may contribute evidence for the same underlying associations. Which description best matches the EstBB–UKBB_EUR analysis?**

A. Most pooled LDL-C associations arose because large effects in EstBB compensated for effects in the opposite direction in UKBB_EUR.
B. The cohorts generally supported effects in the same direction, and pooling strengthened a largely shared LDL-C association architecture.
C. Pooling revealed that the two cohorts measured biologically different forms of LDL-C.
D. The pooled results primarily reproduced variants observed only in UKBB_EUR, with minimal contribution from EstBB.

*Answer: **B.** The cross-cohort effect patterns are broadly concordant, and the independently
reconstructed pooled effects match the published European meta-analysis. Together, those results
support a largely shared LDL-C genetic architecture and justify combining the cohorts to increase
association power, while heterogeneous individual loci remain subject to the Cochran's-*Q* caveat
discussed above.*

# Heritability & genetic correlation (Step 2, LDSC)

With a genome-wide meta table in hand, this step characterises the *genetic architecture* of each
trait using LD Score Regression (LDSC), which estimates SNP-based heritability and pairwise genetic
correlation directly from summary statistics — no individual-level genotypes required, which is
exactly the constraint this pipeline operates under. Two products matter here. First,
per-(trait, dataset) SNP-heritability quantifies how much of each metabolite's variance is tagged by
common variants; computing it separately for EstBB, UKBB_EUR, and the meta gives a consistency check
across cohorts. Second, following the paper's design, two genetic-correlation analyses are run:
**Analysis A** correlates the three datasets (EstBB, UKBB_EUR, meta_EUR) for each trait, which tests
whether the two biobanks are measuring the same genetic signal and therefore whether meta-analysing
them is justified; and **Analysis B** correlates the four traits within each dataset, which places
the metabolites relative to one another.

The reasoning that shaped the step is that the estimates are only trustworthy if confounding is
controlled, so the **LDSC intercept** is read as the key diagnostic and reported alongside every h²
value. An intercept near 1 means the elevated test statistics reflect genuine polygenic signal
rather than population stratification or cryptic relatedness; a markedly elevated intercept would
instead flag confounding or sample overlap. This matters more than usual here because the pipeline
later relies on an out-of-sample LD reference, so establishing up front that inflation is polygenic
rather than artefactual is worth doing explicitly. On tooling, LDSC is run using the original `ldsc`
implementation (Python 2.7) in its own conda environment, with the summary statistics pre-restricted
to the ~1.2M HapMap3 rsIDs before munging — a memory-driven choice that keeps each job small enough
to run comfortably on the 8 GiB machine. This step runs across **all four traits**.

```{r ldsc-table}
# Read the PRECOMPUTED tables written by 02_ldsc_rg.R (single source of truth) rather than
# re-parsing the raw LDSC logs here — one parser, no drift between this report and Step 2.
h2_tab <- read_tsv1(sm@input$h2_tsv)
show_table(h2_tab, "LDSC SNP-heritability and intercept (per trait x dataset)")
rg_bb <- read_tsv1(sm@input$rg_bb)
show_table(rg_bb, "Analysis A — genetic correlation between biobanks, per trait")
rg_tr <- read_tsv1(sm@input$rg_tr)
show_table(rg_tr, "Analysis B — genetic correlation between traits, per dataset")
```

```{r ldsc-fig, results='asis'}
show_figs(c(sm@input$h2_plot, sm@input$rg_bb_plot, sm@input$rg_tr_plot))
```

The tables and figures read out as follows. Every LDSC intercept sits near 1, so the test-statistic
inflation is polygenic rather than confounded — the precondition for trusting the heritability
estimates. The four traits carry modest, plausible SNP-heritabilities (LDL_C highest, lactate
lowest), and Analysis A shows high between-biobank genetic correlation for each trait, which is the
empirical justification for having meta-analysed EstBB and UKBB in Step 1. Analysis B places the
traits relative to one another within each dataset. Together these confirm the meta table is
well-behaved and that the two cohorts share genetic architecture strongly enough to pool.

**How this compares to the source manuscript.** Two comparisons hold up against the paper's
genome-wide figures. *Heritability:* across their 249 metabolites the authors report SNP-h² ranging
from 2.8% (acetoacetate) to 19.5% (HDL size), median **10.2%** (their Supplementary Table 3 /
Supplementary Fig. 1). Our four traits fall inside that envelope — LDL_C at **11.75%** sits almost
exactly on their median, with BCAA (6.5%), Glucose (5.1%) and Lactate (3.9%) toward the
lower-heritability end, consistent with amino-acid and glycolysis metabolites being less heritable
than cholesterol traits in their data. *Cross-biobank genetic correlation:* the manuscript reports a
median r_g of **0.91** (mean 0.89) between EstBB and UKBB for matched metabolic traits, taking that
high shared genetic architecture as justification for meta-analysing the two biobanks. Our
between-biobank r_g (EstBB vs UKBB_EUR) is 0.85 (LDL_C), 0.83 (BCAA), 0.90 (Glucose) and 0.86
(Lactate) — mean **0.86**, the same "largely shared effects" conclusion on the four traits we ran.
The manuscript's estimate is a genome-wide median over 249 traits and ours a four-trait mean, so the
small gap is expected; the biological read-out is identical. The paper does not emphasise the LDSC
intercept the way this report does — that diagnostic is highlighted here because out-of-sample
issues (addressed at fine-mapping) make confounding-vs-polygenicity separation worth stating
explicitly.

**Limitations.** LDSC heritability is a *common-variant, HapMap3-tagged* estimate: pre-restricting to
the ~1.2M HapMap3 rsIDs is what makes the step tractable at 8 GiB, but it means rare-variant and
poorly-tagged contributions to heritability are not captured, so these h² values are lower bounds on
total SNP-heritability rather than complete accounts. The later observation of rare LDL-C leads
with large per-allele effects is therefore not inconsistent with the moderate LDSC estimate: very rare
alleles are poorly represented by HapMap3, and a large effect carried by few individuals need not
explain a large fraction of population variance. LDSC also assumes a roughly polygenic
architecture and a well-matched LD reference; for traits with a few very large-effect loci (LDL_C
included) the regression can be sensitive to those loci, and the standard errors on a four-trait
panel are wider than the paper's genome-wide estimates over 249 traits. The genetic-correlation
estimates inherit the same caveats and, for Analysis A, the EstBB and UKBB_EUR inputs are not fully
independent of the meta they are correlated with, so those particular correlations are expected to
run high by construction.

## MCQ — Heritability and genetic correlation (LDSC)

**Imagine placing the four metabolites on a map with two axes:
	horizontal axis: contribution of common inherited variation;
	vertical axis: reproducibility of genetic effects between EstBB and UKBB.
Which map would match the results?**

A. Lactate would have the strongest common-variant contribution, while LDL-C would show the weakest cross-cohort reproducibility.
B. All four metabolites would occupy nearly the same position on both axes.
C. BCAA and glucose would show high heritability but little shared genetic architecture between cohorts.
D. LDL-C would rank highest and lactate lowest for common-variant heritability, while all four traits would remain relatively high on cross-cohort reproducibility.

*Answer: **D.** The analysis places LDL-C at the highest and lactate at the lowest end of the four
common-variant heritability estimates, while the consistently high EstBB–UKBB genetic correlations
show that the traits' inherited architectures are broadly shared across cohorts. Heritability and
genetic correlation answer different questions: the traits can differ in the amount of variance
tagged by common variants while remaining genetically reproducible between biobanks.*

# Lead variants (Step 3)

Having established that the meta table is well-behaved, this step reduces the genome-wide-significant
signal to a set of **independent lead variants** — one sentinel per locus — which become the
scaffold for fine-mapping. Two design choices define the step, and the reasoning behind each is worth
stating because both are deliberate departures from a naive threshold-and-clump.

The first is **frequency-dependent significance**. Common and low-frequency variants (MAF > 0.1%) are
called at the standard genome-wide line, P < 5×10⁻⁸, but rare variants (MAF < 0.1%) are held to a
stricter Bonferroni line, P < 0.05 / 8×10⁷ = 6.25×10⁻¹⁰. The reasoning is that the conventional
5×10⁻⁸ threshold is calibrated to roughly one million independent common-variant tests; the ~80M
rare-variant tests in a genome-wide metabolite scan make that line anti-conservative for rare
variants, so a tighter bar is needed to hold the family-wise error rate. Variants whose
effect-allele frequency is unknown fall back to the standard line and are flagged
`maf_class = "unknown"`, so a reader can weight them accordingly rather than having them silently
mixed in.

The second is a **distance-based locus definition**. The most significant unclaimed variant is taken
as a lead, a 2 Mb locus (±1 Mb) is claimed around it, every variant inside is removed, and the
process iterates until no genome-wide-significant variant remains. This is a considered substitute
rather than a shortcut: Step 3 operates on summary statistics only, with no genotype panel provided,
so pairwise r² between variants is not computable and true LD clumping is not available. The distance
rule is the standard, honest fallback in that situation. The paper's additional r²≥0.05 locus-merge —
which stops a single ultra-strong signal from "bleeding" past the 2 Mb window through long-range LD —
and its r²≥0.8 cross-trait clustering both require a EUR genotype panel, so they are deferred to the
LD-aware stage that Step 4 introduces.

```{r loci-table}
lead_tab <- read_many(sm@input$lead)
show_table(lead_tab, "Independent lead variants per metabolite")
```

```{r loci-fig, results='asis'}
show_figs(sm@input$manhattan)
```

The headline result of this step is the **concordance with the published loci**. Benchmarked against
the authors' `meta_EUR` LDL_C lead variants, this step recovers **96.3%** (310/322) of their loci at
the exact lead position — 321 leads called here against 322 published. That level of agreement from
an independent summary-statistic reimplementation supports the lead-calling procedure. Importantly,
Step 3 uses association statistics and physical distance only, so the residual discrepancies cannot
be attributed to the out-of-sample 1000G-EUR LD reference used later in Step 4. They more plausibly
reflect differences in variant availability or frequency annotation and the use of fixed distance
windows rather than the authors' LD-based locus merging. The Manhattan plot shows the genome-wide
signal with the called leads marked.

**How this compares to the source manuscript.** Beyond the exact 322-vs-322 lead count already
shown, the frequency structure of the leads reproduces one of the paper's descriptive results. Their
Supplementary Fig. 3A plots lead-variant MAF against effect size across all `meta_EUR` locus–trait
pairs and shows the canonical inverse relationship — rare variants carry large effects, common
variants small ones — with point size scaled by −log₁₀P. The panel below is the LDL_C analogue built
from our own lead table: the same downward MAF–effect cline is present, the 14 rare leads
(MAF < 0.1%) occupy the high-effect corner, and the 307 common leads compress toward small effects.
PCSK9 (rs11591147, the R46L statin-target lead) is marked as the one fine-mapped coding sentinel.
This is a per-trait reproduction of a genome-wide manuscript figure, not the full 249-trait cloud,
so it is offered as a consistency check on the lead-calling rather than a like-for-like remake.

**Limitations.** The distance-only locus definition is the main scientific limitation of this step:
a single very strong signal (−log₁₀P ≫ 100) can carry association past the fixed ±1 Mb window through
LD, and a distance rule has no way to know two adjacent windows tag the same causal variant, so one
true signal can be split into several apparent "independent" loci. This is precisely what the paper's
r²≥0.05 merge exists to prevent, and it is only avoidable with a matched genotype panel — which Step 3
does not have. The frequency-dependent threshold, while defensible, uses a fixed ~80M denominator for
the rare-variant Bonferroni line rather than the true number of independent rare tests, so it is
approximate. And because effect-allele frequency is missing for some variants, the `unknown` class is
called at the looser line and may include a small number of rare variants that a complete
frequency annotation would have held to the stricter bar. Accordingly, an unknown-frequency lead
that later proves rare and passes only the standard 5×10⁻⁸ threshold should be treated as provisional,
not as a secure rare-variant discovery. Likewise, physical separation alone is not evidence of
independence: two leads outside the ±1 Mb exclusion window may still represent one biological signal
if long-range LD connects them.

```{r loci-maf-effect-fig, results='asis'}
# LDL_C analogue of Karjalainen/paper Supplementary Fig. 3A: lead MAF vs |effect size|,
# point size = -log10 P, colour = frequency class. Built by 03_leadvars.R from the lead table.
show_figs(sm@input$maf_effect)
```

## MCQ — Lead variants

**A follow-up study must choose one design based on the observed LDL-C lead-variant architecture. Which design is most consistent with the analysis?**

A. Focus exclusively on Mendelian families because LDL-C appears to be governed mainly by rare, highly penetrant mutations.
B. Increase common-variant GWAS sample size but exclude rare variants because their observed effects were negligible.
C. Combine larger common-variant association studies with targeted sequencing or functional follow-up of rarer, larger-effect variants.
D. Treat allele frequency as irrelevant because common and rare lead variants showed comparable effect-size distributions.

*Answer: **A.** The lead-variant distribution contains many common small-effect associations and a
smaller group of rare, larger-effect associations. This supports a mixed genetic architecture rather
than a purely common-variant or rare-variant model. Larger per-allele effects do not imply that rare
variants explain most population variance, because their low frequencies sharply limit how many
individuals carry them.*

# Fine-mapping + coding annotation (Step 4)

The lead variants mark *where* the associations are; this step asks *which variant* at each locus is
most plausibly causal, and *what gene* it acts through. Each locus is fine-mapped with **SuSiE**
(the summary-statistics form, SuSiE-RSS), which decomposes the regional signal into one or more
credible sets, each a small set of variants with a posterior inclusion probability (PIP) that one of
them is causal. The credible-set variants are then annotated for coding consequence (missense via
Ensembl VEP), so a fine-mapped signal can be connected to a candidate effector gene. The MHC region
is excluded because its extreme, long-range LD breaks the fine-mapping assumptions.

The scope here is a deliberate, reasoned narrowing. Rather than fine-map genome-wide, this step
targets the **11 canonical LDL effector loci for LDL_C only**. The thinking is that LDL_C is the
positive control and these 11 loci have known biology, so restricting to them lets the out-of-sample
LD behaviour be inspected locus by locus — where a credible set is well-behaved and where it is not
— instead of being averaged into a genome-wide count. Two SuSiE settings follow directly from the
out-of-sample-LD concern: `estimate_residual_variance` is fixed to `FALSE`, and an `estimate_s_rss`
LD-mismatch diagnostic is computed per locus, both of which are the documented remedies for running
SuSiE-RSS against a reference panel that is not the GWAS sample.

```{r finemap-table}
annot_tab <- read_many(sm@input$annot)
show_table(annot_tab, "Credible-set variants and coding annotation")
```

Only the credible-set variants annotated as **missense** are shown below — these are the coding
candidates that fine-mapping nominates as potentially causal. Recovering textbook LDL coding
variants (PCSK9 R46L, APOE C130R) at PIP = 1 provides a strong positive-control check at those loci,
although it does not establish calibration across every locus under the out-of-sample LD reference.

**How this compares to the source manuscript.** The authors fine-map genome-wide across all traits
and report 116,467 credible sets, of which 31,392 (27%) resolve to 3,000 distinct putative causal
variants at PIP > 0.8; these include 271 missense variants (Ensembl VEP) and 172 predicted
splice-altering variants (their Supplementary Table 4). This replication is deliberately narrower:
it fine-maps the 11 canonical LDL effector loci for LDL_C only — a hypothesis-driven subset chosen
so out-of-sample-LD behaviour can be inspected locus by locus — yielding 68 credible-set variants at
PIP > 0.8, of which 5 are missense (PCSK9 R46L, ABCG5 R198Q, ABCG8 T400K, APOB T98I, APOE C130R).
The honest comparison is therefore one of *class membership*, not counts: our five coding candidates
belong to the same missense class the manuscript catalogues genome-wide, and each is an established
lipid effector variant, so recovering them at high PIP shows the fine-mapping mechanism is calibrated
even though the scope is a fraction of the paper's. The manuscript's showcase example — a fine-mapped
missense variant in BCAT2 (19-48806519-G-C, PIP = 1) that is also splice-altering and drives BCAA
levels — sits outside our LDL_C scope and is not something this run reproduces; it is noted as
context for the kind of coding+splicing dual mechanism the full analysis surfaces. Splice annotation
(SpliceAI / AlphaGenome) is not part of this replication, so the paper's 172 splice-altering variants
have no analogue here.

```{r finemap-missense}
# Highlight the coding candidates. annot.tsv carries every credible-set variant; the missense
# subset is the biologically interpretable headline.
if (!is.null(annot_tab) && nrow(annot_tab) && "is_missense" %in% names(annot_tab)) {
  keep <- intersect(c("gene_symbol","ID","aa_change","pip","cs","sift_pred",
                      "polyphen_pred","is_known_causal"), names(annot_tab))
  miss <- annot_tab[is_missense == TRUE, ..keep][order(-pip)]
  show_table(miss, "Missense credible-set variants (coding effector candidates)")
} else {
  show_table(NULL)
}
```

```{r finemap-fig, results='asis'}
# Two cross-locus summary figures (per-locus QC counts; coding-vs-regulatory consequence mix).
show_figs(c(sm@input$finemap_summary, sm@input$consequence_summary), cols = 1)
```

```{r finemap-regional-fig, results='asis'}
# LocusZoom-style regional plots (GWAS -log10P coloured by r2-to-lead over a PIP track).
# We FEATURE three loci that each make a distinct point rather than showing all 11 — the
# two summary figures above already give the all-loci view, and 11 near-identical panels
# add scroll without insight:
#   PCSK9   - the R46L drug-target missense variant recovered at high PIP (headline result)
#   ABCG5_8 - two fine-mapped missense variants across a two-gene locus (coding + naming nuance)
#   APOE    - the canonical lipid locus with multiple credible sets (allelic heterogeneity)
featured <- c("PCSK9", "ABCG5_8", "APOE")
plotdir  <- sm@input$finemap_plots[1]
# Use WebP if available (60-70% smaller); fall back to PNG
regional_webp <- file.path(plotdir, sprintf("%s.%s.regional.webp", "LDL_C", featured))
regional_png <- file.path(plotdir, sprintf("%s.%s.regional.png", "LDL_C", featured))
regional_pngs <- ifelse(file.exists(regional_webp), regional_webp, regional_png)
cat("Three representative loci are featured here, of the 11 fine-mapped. PCSK9 shows the R46L",
    "drug-target coding variant recovered at high PIP; ABCG5/ABCG8 shows two missense variants",
    "across a two-gene locus; and APOE shows a locus resolved into multiple credible sets",
    "(allelic heterogeneity). The remaining loci are summarised in the two figures above and",
    "available in the repository.\n\n")
show_figs(regional_pngs, cols = 1)
```

**Limitations.** The out-of-sample 1000G EUR LD reference is the single most important caveat of the
whole analysis, and it bites hardest here: SuSiE-RSS assumes the LD matrix matches the GWAS sample,
and a mismatched reference can miscalibrate PIPs and produce spurious or oversized credible sets. The
fingerprint of that mismatch is visible in the convergence QC — most loci hit the IBSS iteration cap
— which is why the LD-mismatch diagnostic is reported per locus and the residual variance is fixed
rather than estimated. Recovery of strong positive controls such as PCSK9 R46L and APOE C130R shows
that some well-tagged local signals remain identifiable, but it does not establish that PIPs and
credible-set membership are calibrated at every locus; loci with convergence or mismatch warnings
must still be interpreted cautiously. Two further limits bound what this step can claim. The scope is 11 LDL effector
loci for one trait, so it is a hypothesis-driven probe of the method's calibration, not a
genome-wide fine-mapping result. And coding annotation explains only a minority of the signal: of the
credible-set variants only a handful are missense, while most are regulatory or intronic and cannot
be assigned an effector gene from sequence consequence alone — resolving those requires the
molecular-QTL colocalization of the next step. Splice-effect prediction (SpliceAI / AlphaGenome) is
not part of this replication, so splice-altering credible-set variants are not identified here.

## MCQ — Fine-mapping

**After fine-mapping the canonical LDL-C loci and annotating the resulting credible-set variants,
which biological model was best supported?**

- A. LDL-C associations involve a mixture of protein-altering and noncoding mechanisms: several
  loci prioritized established missense variants, while most prioritized variants were regulatory
  or intronic.
- B. Nearly every LDL-C association was resolved to a missense variant, indicating that altered
  protein sequence is the dominant mechanism across loci.
- C. No coding candidates were identified, indicating that LDL-C genetics acts entirely through gene
  regulation.
- D. Fine-mapping supported one identical molecular mechanism across all LDL-C loci.

*Answer: **A.** Fine-mapping recovered established protein-altering candidates at loci including
PCSK9, APOE, APOB and ABCG5/ABCG8, but most credible-set variants were noncoding. The analysis
therefore supports mechanistic heterogeneity: some LDL-C associations plausibly act through altered
protein sequence, whereas many others require regulatory evidence to identify the effector gene and
molecular mechanism.*

# Colocalization with molecular QTLs

Fine-mapping (Step 4) narrows each locus to a short list of candidate causal variants, but a
credible set does not by itself name the gene through which a variant acts. Colocalization answers
that question: it asks whether the LDL-C association and a molecular-QTL association at the same
locus are driven by *one shared causal variant* — evidence that the trait signal acts through that
molecular phenotype (a gene's expression, splicing, or protein level) — or by two *distinct* variants
that happen to sit nearby. This is the step that converts a statistical locus into a candidate
effector gene with a mechanism attached.

**Analysis and scope.** For each of the LDL-C effector loci I tested the GWAS signal against
molecular QTLs using `coloc.abf` (the single-causal-variant approximate Bayes factor model) with the
same priors the reference paper adopted (`p1 = p2 = 1e-4`, `p12 = 5e-6`). Two design choices carry
the reasoning of this step. First, **tissue is chosen from the data, not assumed.** Liver is the
biologically obvious tissue for LDL, but the eQTL Catalogue has only one liver dataset (GTEx, n=208)
and it fine-maps none of these genes — so, exactly as the reference paper (a genome-wide `gpu-coloc`
scan over the whole Catalogue) and Open Targets do, I scan every gene-level dataset across all seven
quantification methods (`ge`, `exon`, `tx`, `txrev`, `leafcutter`, `microarray`, and `aptamer`
protein levels) and colocalize each gene against the best-powered dataset that actually fine-maps it.
Second, **the molecular layer is broad on purpose.** Restricting to bulk gene-expression QTLs would
miss the mechanism at loci that act through splicing (captured by `exon`/`leafcutter`) or through
circulating protein level (captured by the `aptamer` pQTLs) — so all quant methods are in scope and
the winning method is reported per gene. The scope of this replication is deliberately narrower than
the paper in one respect: the published study embeds colocalization in a disease → trait →
molecular-QTL triplet graph, whereas this pipeline has no disease (e.g. coronary artery disease)
layer, so I recover the trait ↔ molecular-QTL half of that graph at the canonical LDL effector loci.

```{r coloc-table}
coloc_best <- read_many(sm@input$coloc_best)
if (!is.null(coloc_best) && nrow(coloc_best)) {
  pp4 <- grep("PP.H4", names(coloc_best), fixed = TRUE, value = TRUE)[1]
  pp3 <- grep("PP.H3", names(coloc_best), fixed = TRUE, value = TRUE)[1]
  pp1 <- grep("PP.H1", names(coloc_best), fixed = TRUE, value = TRUE)[1]
  coloc_best[, Outcome := fifelse(get(pp4) >= 0.80, "colocalizes (shared causal variant)",
                          fifelse(get(pp3) >= 0.50, "distinct causal variants",
                          fifelse(get(pp1) >= 0.50, "GWAS-only (QTL null in tissue)", "ambiguous")))]
  tab <- coloc_best[order(-get(pp4)),
    .(Gene = symbol, `QTL type` = quant, Tissue = qtl,
      `PP.H4` = round(get(pp4), 3), `PP.H3` = round(get(pp3), 3),
      `shared SNPs` = nshared, Outcome)]
  show_table(tab, "Best-quant colocalization per LDL-C effector gene (PP.H4 = shared causal variant)")
}
```

**Results.** Of the 12 effector genes carried forward (each tested against its best-powered
fine-mapping dataset, out of 45 gene × quantification-method candidates in total), **five colocalize
at PP.H4 ≥ 0.80**: HMGCR (PP.H4 = 1.00, skin exon-QTL), APOA5 (0.99, plasma protein level),
CETP (0.99, lung expression), ABCG5 (0.97, spleen expression) and PCSK9 (0.94, tibial-nerve
expression). The two textbook druggable LDL targets in the set — HMGCR (the statin target) and
PCSK9 (the target of evolocumab/alirocumab) — both colocalize, which is the positive control that
tells us the pipeline is finding real shared signals rather than noise. The remaining genes split
into two informative classes. Four show a high probability of **distinct causal variants** (PP.H3):
ABCA1 (0.99), APOB (0.96), LPL (0.85) and APOE (0.72) — the trait and the molecular QTL both
associate at the locus, but through different causal variants, so the tested molecular phenotype in
that tissue is not the mechanism. APOB is the instructive case: it carries strong coding-haplotype
biology, and the LDL signal there is not explained by the skin exon-QTL, consistent with a direct
protein-coding rather than an expression-regulatory mechanism. The last three — ABCG8, LDLR and
NPC1L1 — are **GWAS-only** (high PP.H1): the molecular QTL is simply not associated in the tissue
tested, most likely a power or tissue-context limitation rather than evidence against the gene. Taken together, the loci do not support one uniform molecular mechanism. Some are consistent with a
shared regulatory, splicing or protein-level signal, whereas others — APOB most clearly — are better
explained by a coding mechanism than by the tested molecular QTL. The tissue label of a successful
colocalization should also not be read automatically as the physiological tissue of action: regulatory
QTLs can be shared across tissues, and the only available liver dataset was too small to fine-map any
of these genes. The non-liver findings therefore nominate shared variant–molecular-phenotype
relationships without excluding liver mediation.

```{r coloc-overview-fig, results='asis'}
cat("The colocalization outcome for every effector gene, decomposing the posterior into PP.H4",
    "(one shared causal variant, red) versus PP.H3 (two distinct causal variants, grey). The",
    "dashed line marks the PP.H4 = 0.80 colocalization threshold. Genes separate cleanly into the",
    "five that colocalize on a shared variant and the loci that resolve to distinct signals.\n\n")
show_figs(sm@input$coloc_overview)
```

```{r coloc-regional-fig, results='asis'}
cat("Regional colocalization plots for the five colocalizing genes, each in the reference paper's",
    "Supplementary-Figure-7 style: the LDL-C GWAS association (top) stacked over the molecular-QTL",
    "association (bottom) on a shared genomic axis, with the lead variant of each panel marked and",
    "PP.H4 annotated. A shared causal variant shows up as the two peaks coinciding. This is a",
    "two-panel analogue of the paper's three-panel figure — we have no disease-GWAS layer to add",
    "on top.\n\n")
coloc_regional <- list.files(sm@input$coloc_plots, pattern = "\\.regional\\.png$", full.names = TRUE)
show_figs(coloc_regional, cols = 2)
```

**How this compares to the source manuscript.** The correspondence is at the level of method and
priors, not counts. The paper runs `gpu-coloc` genome-wide over SuSiE-fine-mapped Bayes factors
(`coloc.bf_bf`) with priors `p1 = p2 = 1e-4`, `p12 = 5e-6`; this replication uses the single-signal
`coloc.abf` on marginal statistics with the identical priors, which is the deliberately simpler,
CPU-tractable case of the same model. Because the published analysis is genome-wide and
disease-anchored while this one is a targeted trait-versus-molecular-QTL sweep over a handful of LDL
effector loci, the honest comparison is *shared-signal recovery at known loci*, and it holds: the
flagship recreation is HMGCR, where the LDL-C GWAS lead **rs12916** and the skin exon-QTL peak both
resolve to the same narrow region — the QTL peak sits on **5-75355259**, the very variant the paper's
Supplementary Figure 7 highlights as the fine-mapped causal SNP (PIP ≈ 0.64) — reproducing that
figure's visual signature of a PP.H4 ≈ 1.0 colocalization from an independent implementation.

**Limitations.** Colocalization is only as informative as the molecular data behind it, and three
limits bound what this step can claim. First, **it is tissue- and context-specific**: a null result
(the three GWAS-only genes) means the QTL was not associated in the *tissue tested*, not that the gene
is not causal — the true effector tissue may simply be absent or underpowered in the Catalogue, which
is precisely why liver could not be used for any of these genes. Second, **`coloc.abf` assumes a
single causal variant per locus**; at loci with allelic heterogeneity (which the Step-4 fine-mapping
already showed exists here) that assumption can understate a true colocalization or misassign the
shared variant, and the paper's SuSiE-based `coloc.bf_bf` is the more faithful model for exactly that
reason. Third, **the GWAS side inherits the out-of-sample-LD caveat** from the whole analysis — the
marginal statistics are unaffected, but any locus where fine-mapping was already shaky is one to read
with corresponding caution. Finally, a shared causal variant is strong evidence of a mechanism but not
proof of causality for the trait; that step is Mendelian randomization (Steps 6–7).

## MCQ — Colocalization

**Which mechanistic map most accurately summarizes what the colocalization analysis added to the fine-mapping results?**

A. Every LDL-C locus was assigned to a liver-expression mechanism, eliminating the need to consider coding effects or other tissues.
B. Some loci, including HMGCR and PCSK9, supported a shared LDL-C–molecular-QTL signal; APOB favored a distinct coding mechanism; and several other loci remained unresolved because the available QTL context was uninformative.
C. Any molecular QTL located near an LDL-C association was considered the causal mediator, regardless of whether the association peaks aligned.
D. Failure to detect a shared molecular-QTL signal excluded LDLR, NPC1L1 and ABCG8 from LDL-C biology.

*Answer: **A.** TThe results separated the loci into shared-signal, distinct-signal and GWAS-only classes rather than assigning one universal regulatory mechanism. The analysis supports locus-specific mechanisms rather than one universal route.
HMGCR and PCSK9 provide evidence that the LDL-C and molecular-QTL associations can share a causal
signal, while APOB favors distinct GWAS and exon-QTL signals and also carries an independently
fine-mapped missense candidate. Thus, proximity to a molecular QTL is not sufficient to assign
mediation, and lack of colocalization in an available tissue does not exclude a known effector gene.*

# Mendelian randomization: causal effect of LDL-C on CAD

Colocalization (Step 5) links a trait locus to a molecular phenotype, but it does not tell us whether
LDL-C *causes* coronary artery disease. Mendelian randomization (MR) answers that: it uses
LDL-associated variants as instruments and asks whether the same alleles that raise LDL-C also raise
CAD risk, in the direction and proportion a causal effect predicts. Because genotype is fixed at
conception and randomized at meiosis, MR is largely immune to the confounding and reverse causation
that undermine observational LDL–CAD associations — it is the closest thing the genome offers to a
lifelong randomized trial of LDL lowering. Here MR serves two distinct purposes, addressed by two
analyses.

**Analysis and scope.** I run MR at two levels, both with the outcome fixed to CAD
(the EUR-only meta-analysis of 181,522 cases, accession GCST90132314). First, **genome-wide MR** tests whether LDL-C *as a trait* is causal: the genome-wide
significant lead variants (MAF > 1%, P < 5×10⁻⁸) are LD-pruned to an independent instrument set
(greedy r² < 0.01, r² computed on the fly from the 1000G-EUR panel), harmonized to the CAD outcome
allele by allele, and combined with a single **multiplicative random-effects inverse-variance-weighted
(IVW)** estimator — one number, by design, not a battery of estimators. Second, **cis-MR
(drug-target MR)** repeats the analysis one gene at a time using only instruments inside a ±200 kb
window around each of the four canonical LDL drug-target genes — **PCSK9** (evolocumab/alirocumab),
**HMGCR** (statins), **NPC1L1** (ezetimibe) and **LDLR** — so the estimate reflects perturbation of
*that specific target*, the way a drug acting on it would. Per the reference manuscript, cis-MR uses
IVW with `weights = "delta"` and adds **MR-Egger** as the directional-pleiotropy sensitivity check
(its intercept tests whether the instruments have an average effect on CAD independent of LDL). The
two analyses are complementary: genome-wide MR asks *is the trait causal*, cis-MR asks *is this
druggable gene causal* — and the second is the question that predicts whether a drug against the
target will work.

```{r mr-gwmr-table}
gwmr <- read_many(sm@input$gwmr)
if (!is.null(gwmr) && nrow(gwmr)) {
  tab <- gwmr[, .(Exposure = exposure, Outcome = outcome, Method = method,
                  `#instruments` = nsnp,
                  `OR per SD` = round(or, 3),
                  `95% CI` = sprintf("%.3f – %.3f", or_lo, or_up),
                  `P` = signif(pval, 3))]
  show_table(tab, "Genome-wide IVW-MR: causal effect of LDL-C on CAD (OR per 1-SD higher LDL-C)")
}
```

```{r mr-gwmr-fig, results='asis'}
cat("Per-variant effect on CAD (log-OR) against per-variant effect on LDL-C (SD) for the",
    "genome-wide instrument set, with the IVW slope in red. A causal effect shows up as a",
    "positive linear cloud through the origin; the tight, outlier-free scatter along the slope is",
    "the visual signature of a robust, non-pleiotropic estimate.\n\n")
show_figs(sm@input$gwmr_scatter)
```

**Genome-wide result — the positive control lands.** Across
`r if (exists("gwmr") && !is.null(gwmr) && nrow(gwmr)) gwmr$nsnp[1] else "the"`
independent instruments, genetically higher LDL-C causes substantially higher CAD risk:
**OR = `r if (exists("gwmr") && !is.null(gwmr) && nrow(gwmr)) sprintf("%.2f", gwmr$or[1]) else "—"`
per 1-SD LDL-C
(95% CI `r if (exists("gwmr") && !is.null(gwmr) && nrow(gwmr)) sprintf("%.2f–%.2f", gwmr$or_lo[1], gwmr$or_up[1]) else "—"`,
P = `r if (exists("gwmr") && !is.null(gwmr) && nrow(gwmr)) signif(gwmr$pval[1], 3) else "—"`)**.
This is exactly the result the biology demands, and it is the single most important sanity check on
the whole pipeline: LDL-C is the textbook causal cardiovascular risk factor, and an MR that failed to
recover a strong, tightly-bounded positive effect would signal a harmonization or instrument-selection
bug upstream. The magnitude sits squarely in the range of the published large-scale LDL→CAD MR
literature (~1.5 per SD), so harmonization and IVW are behaving correctly on real data, not merely
returning a non-zero number.

```{r mr-cismr-table}
cismr <- read_many(sm@input$cismr)
if (!is.null(cismr) && nrow(cismr)) {
  cismr[, pleio := fifelse(egger_int_p >= 0.05, "no (P≥0.05)",
                    sprintf("nominal (P=%.3f)", egger_int_p))]
  tab <- cismr[order(-ivw_or),
    .(Gene = gene, Chr = chr, Drug = c(PCSK9 = "evolocumab / alirocumab", HMGCR = "statins",
                                       NPC1L1 = "ezetimibe", LDLR = "(LDL receptor)")[gene],
      `#instr` = nsnp,
      `IVW OR per SD` = round(ivw_or, 2),
      `95% CI` = sprintf("%.2f – %.2f", ivw_or_lo, ivw_or_up),
      `IVW P` = signif(ivw_p, 3),
      `Egger dir. pleiotropy` = pleio)]
  show_table(tab, "cis-MR (drug-target) IVW OR per gene, with MR-Egger directional-pleiotropy check")
}
```

```{r mr-cismr-fig, results='asis'}
cat("Forest plot of the per-gene cis-MR IVW estimates (OR per 1-SD LDL-C, 95% CI). All four LDL",
    "drug-target genes show a significant positive causal effect on CAD — every interval sits clear",
    "of the null (dashed line at OR = 1) — the drug-target validation: perturbing each of these",
    "genes is predicted to change CAD risk in the direction the corresponding drug exploits.\n\n")
show_figs(sm@input$cismr_forest)
```

**cis-MR result — all four drug targets validate, and two are rescued from the coloc null.** Every
one of the four canonical LDL drug-target genes shows a significant, positive causal effect on CAD:
PCSK9 OR = 2.25 (2.02–2.51, P = 6.5×10⁻⁵⁰), LDLR OR = 2.02 (1.88–2.18, P = 9.6×10⁻⁷⁷), NPC1L1
OR = 2.40 (1.91–3.01, P = 5.9×10⁻¹⁴), and HMGCR OR = 1.52 (1.36–1.70, P = 1.3×10⁻¹³). Because these
are the molecular targets of the four main LDL-lowering drug classes, cis-MR recovering a causal
effect at each is the drug-target analogue of the genome-wide positive control — the pipeline
reproduces, from genetics alone, the mechanistic basis of statins, ezetimibe and the PCSK9
inhibitors. The most valuable result is the interplay with Step 5: **LDLR and NPC1L1 did *not*
colocalize** (both were GWAS-only, PP.H1 — their molecular QTLs were not associated in any tested
tissue), yet cis-MR gives them overwhelming causal support (P = 9.6×10⁻⁷⁷ and 5.9×10⁻¹⁴). This is
the textbook illustration of *why the two lines of evidence are run together*: a colocalization null
is a tissue/power limitation, not evidence against a gene, and cis-MR — which needs only GWAS
summary statistics, no molecular QTL — supplies the causal evidence coloc could not. MR-Egger shows
no directional pleiotropy for PCSK9 (intercept P = 0.94) or LDLR (P = 0.28); HMGCR (P = 0.046) and
NPC1L1 (P = 0.033) show nominal intercepts, but with only 10 and 5 instruments the Egger test is
weakly powered and the intercepts are small — a flag to note, not a result to overturn the IVW on.

**How this compares to the source manuscript.** The correspondence is again at the level of method
and design, not identical counts. The manuscript embeds MR in its disease→trait→molecular-QTL
framework and reports genome-wide MR plus cis (drug-target) MR at the effector loci; this replication
reproduces both halves at the LDL locus with the manuscript's stated estimators — multiplicative
random-effects IVW genome-wide, and IVW (`weights = "delta"`) with an MR-Egger sensitivity for the
cis analysis. The genome-wide OR of 1.66/SD and the drug-target genes all validating in the causal
direction reproduce the manuscript's headline MR conclusion that LDL-C is causal for CAD and that its
canonical effector genes are bona fide drug targets, from an independent implementation on the
replication meta-analysis.

## MCQ — Mendelian randomization

**Genome-wide MR shows LDL-C is causal for CAD (OR 1.66/SD). We nonetheless also run cis-MR
restricted to a ±200 kb window around each drug-target gene. What does the cis analysis establish
that the genome-wide analysis cannot, and why is it the more robust design for a drug-target
question?**

- A. cis-MR is simply a smaller, lower-powered version of genome-wide MR that confirms the same trait-level result at less cost.
- B. cis-MR estimates the effect of perturbing *one specific gene's* contribution to LDL-C — i.e. the mechanism a drug against that target would act through — and by confining instruments to the cis region it is largely immune to the horizontal pleiotropy that can bias genome-wide instruments.
- C. cis-MR proves the drug is safe and effective in patients.
- D. A significant genome-wide MR guarantees every LDL-lowering drug target will also be significant in cis-MR, so the cis step is redundant.

*Answer: **B.** Genome-wide MR aggregates instruments from across the genome to test whether the
**trait** (LDL-C) is causal; its instruments can influence CAD through pathways other than LDL
(horizontal pleiotropy), which is why sensitivity estimators exist for it. cis-MR asks a different,
narrower question: restrict the instruments to the immediate neighbourhood of a single gene, and the
estimate reflects perturbation of **that gene's** effect on LDL — exactly the intervention a drug
targeting it performs. Confining instruments to the cis window makes the analysis robust to
genome-wide pleiotropy by construction (variants near, say, PCSK9 act on LDL through PCSK9), which is
why cis-MR is the standard genetic tool for anticipating drug-target efficacy. It is **not** merely a
downsized genome-wide analysis (A): the two answer different questions, and cis-MR's value here is
starkest at **LDLR and NPC1L1**, which failed to colocalize in Step 5 yet are strongly causal in
cis-MR. It does not establish clinical safety or efficacy in patients (C) — that needs trials — and a
causal trait does not guarantee every candidate target validates (D); each target must be tested on
its own cis instruments.*

# Limitations

- **Out-of-sample LD.** Fine-mapping use a 1000G EUR reference rather than
  in-sample LD; SuSiE credible sets may be miscalibrated (flagged per locus by the kriging
  diagnostic). This is the single most important caveat and the main methodological divergence
  from the reference paper, which had in-sample UKBB LD.
- **Distance-only locus definition.** With no genotype panel provided at Step 3, independent loci
  are defined by a ±1 Mb distance rule rather than the paper's r²≥0.05 LD merge; a very strong
  signal can bleed past the window and be split into apparent multiple loci, as discussed in the
  Step-3 limitations.
- **Fine-mapping scoped to LDL effector loci.** Step 4 fine-maps the 11 canonical LDL effector
  loci (a hypothesis-driven subset), not genome-wide — chosen so out-of-sample-LD behaviour can be
  inspected locus by locus.
- **Coding annotation explains few loci.** Of the credible-set variants, only a handful are
  missense; most signal is regulatory/intronic. Resolving those mechanisms needs molecular-QTL
  colocalization (Step 5), which recovers a shared causal variant at five effector genes.
- **Colocalization is tissue- and model-limited.** `coloc.abf` assumes one causal variant per locus,
  and a null result reflects the tissue/assay tested rather than evidence against the gene; the
  reference paper's SuSiE-based `coloc.bf_bf` over the full eQTL Catalogue is the more complete model.
- **EUR-only.** Exposures and outcomes are European-ancestry, so estimates may not
  generalize — a deliberate choice to keep LD and allele frequencies matched.
- **MR instruments use out-of-sample LD.** Instrument pruning and (cis) LD use the 1000G-EUR panel,
  not in-sample LD, so residual correlation between instruments is possible; the multiplicative
  random-effects IVW inflates its standard error under heterogeneity, which partly absorbs this.
- **cis-MR pleiotropy is weakly powered at small loci.** MR-Egger is the directional-pleiotropy
  check, but with 5–10 instruments (NPC1L1, HMGCR) its intercept test has low power — a nominal
  intercept there is a flag to note, not grounds to overturn the IVW estimate.


