---
title: "From GWAS Summary Statistics to Candidate Causal Genes: A Reproducible Post-GWAS Pipeline for Circulating Metabolic Traits in European-Ancestry Biobanks"
subtitle: "Meta-analysis, heritability, fine-mapping, colocalization, and Mendelian randomization for LDL cholesterol, branched-chain amino acids, glucose, and lactate"
author: "Khushi Goda"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
library(data.table); library(ggplot2); library(DT)
library(cowplot); library(png)   # compose_figs(): tile per-trait PNGs into one image
sm <- snakemake   # S4 object; inputs wired by the `report` rule in the Snakefile
set.seed(sm@params$seed)   # reproducibility: global seed from config

# ---- helpers (make tomorrow "plug data in and knit") ----------------------
# Read one tsv defensively: empty/absent -> NULL (so scaffold runs never error).
read_tsv1 <- function(path) tryCatch(fread(path), error = function(e) NULL)

# Read several tsvs and stack them, tagging each row with its source file.
read_many <- function(paths) {
  rbindlist(lapply(paths, function(p) {
    d <- read_tsv1(p)
    if (!is.null(d) && nrow(d)) d[, source_file := basename(p)]
    d
  }), fill = TRUE, use.names = TRUE)
}

# Render a table, or a placeholder line if there are no rows yet.
show_table <- function(dt, caption = NULL) {
  if (is.null(dt) || nrow(dt) == 0) {
    cat("*No rows yet — this table populates when the pipeline runs on real data.*\n\n")
    return(invisible(NULL))
  }
  DT::datatable(dt, caption = caption, rownames = FALSE,
                options = list(pageLength = 10, scrollX = TRUE))
}

# Embed figures that actually exist (skips scaffold-empty paths cleanly).
# NB: we DON'T use include_graphics here. Snakemake renders the Rmd from a temp
# script dir (.snakemake/scripts/), and both include_graphics and Pandoc's
# self_contained pass resolve image paths relative to THAT dir — so any path we pass
# (relative -> "Could not fetch resource"; absolute -> relativised to "../../results/..")
# fails. Instead we read each file (cwd = knit_root_dir = repo root, so file.exists
# passes) and inline it as a base64 data-URI <img>, which needs no path resolution at
# all. Chunks calling this MUST set results='asis' so the HTML is emitted raw.
show_figs <- function(paths, cols = NULL) {
  paths <- paths[file.exists(paths) & file.info(paths)$size > 0]
  if (!length(paths)) return(invisible(NULL))
  # A single figure renders full-width & centred; multiple figures render in a
  # responsive CSS grid (default 2 cols, auto-fit down to 1 on narrow screens) so
  # groups like the 11 regional LocusZoom panels read as a compact gallery instead
  # of an endless vertical scroll.
  if (is.null(cols)) cols <- if (length(paths) > 1) 2 else 1
  if (cols <= 1 || length(paths) == 1) {
    for (p in paths)
      cat(sprintf('<img src="%s" style="max-width:100%%;display:block;margin:0.5em auto;" alt="%s"/>\n\n',
                  knitr::image_uri(p), basename(p)))
  } else {
    cat(sprintf('<div style="display:grid;grid-template-columns:repeat(%d,1fr);gap:0.6em;margin:0.5em 0;">\n', cols))
    for (p in paths)
      cat(sprintf('<img src="%s" style="width:100%%;height:auto;" alt="%s"/>\n',
                  knitr::image_uri(p), basename(p)))
    cat('</div>\n\n')
  }
  invisible(NULL)
}

# Tile several PNGs into ONE composite image (cowplot) and inline it base64.
# Used for the Step-1 per-trait scatter families (4 validation, 4 concordance):
# one 2x2 image reads far better than 8 separate full-width plots and cuts weight.
# Each source PNG already carries its own trait title, so no cowplot labels.
compose_figs <- function(paths, ncol = 2, panel_in = 4.5) {
  paths <- paths[file.exists(paths) & file.info(paths)$size > 0]
  if (!length(paths)) return(invisible(NULL))
  if (length(paths) == 1) return(show_figs(paths))
  grobs <- lapply(paths, function(p)
    ggdraw() + draw_grob(grid::rasterGrob(readPNG(p), interpolate = TRUE)))
  g <- plot_grid(plotlist = grobs, ncol = ncol)
  nrow <- ceiling(length(paths) / ncol)
  tmp <- tempfile(fileext = ".png")
  ggsave(tmp, g, width = panel_in * ncol, height = panel_in * nrow,
         dpi = 150, device = grDevices::png, type = "cairo", bg = "white")
  cat(sprintf('<img src="%s" style="max-width:100%%;display:block;margin:0.5em auto;" alt="composite"/>\n\n',
              knitr::image_uri(tmp)))
  invisible(NULL)
}
```

# Overview & method choices

> **Reproducibility.** This document is the R Markdown source `workflow/scripts/08_report.Rmd`,
> knitted by the Snakemake `report` rule. Every number, table, and figure below is generated
> at knit time by reading the pipeline's output files in `results/` — nothing is hand-entered
> or pasted. Re-running the pipeline and re-knitting regenerates this report end to end, so the
> report cannot drift from the analysis it describes.

> **Scope of this knit.** This version reports the **completed steps: meta-analysis, LDSC,
> lead-variant definition, and fine-mapping + coding annotation (Steps 1–4)**, each followed by a
> multiple-choice question testing the biological reasoning behind that step. Steps 1–2 run across
> all four traits; Steps 3–4 are shown for **LDL_C**, the positive-control trait whose 11 canonical
> effector loci let the out-of-sample-LD behaviour be inspected locus by locus. Colocalization with
> molecular QTLs (Step 5) is also shown for **LDL_C**; Mendelian randomization (Steps 6–7) is in
> progress and will be folded into this same report — each step with its own MCQ — once its pipeline
> rules have run.

This report follows four circulating metabolic traits — **LDL cholesterol (LDL_C)**, **total
branched-chain amino acids (BCAA)**, **glucose**, and **lactate** — from published GWAS summary
statistics through a standard post-GWAS causal-inference chain: cross-cohort meta-analysis,
heritability and genetic correlation (LDSC), lead-variant definition, statistical fine-mapping,
colocalization with gene expression, and Mendelian randomization against two cardiometabolic
diseases. The analysis is **European-ancestry throughout** (EstBB + UKBB_EUR exposures,
EUR-matched CAD/T2D outcomes, 1000G EUR LD reference), and **no individual-level genotype data
is used** — every step operates on summary statistics plus a public LD reference.

**Data provenance.** All exposure summary statistics come from Tambets *et al.*[^tambets], a GWAS
meta-analysis of 249 circulating metabolic traits in the **Estonian Biobank (EstBB)** and the
**UK Biobank (UKBB)**, retrieved as GWAS-SSF-formatted files from the **NHGRI-EBI GWAS Catalog**.
For each metabolite the pipeline uses the two per-cohort European releases (EstBB and UKBB_EUR) as
meta-analysis inputs and the authors' own published European meta-analysis as the Step-1 validation
target. The **coronary artery disease (CAD)** outcome used downstream is the European-ancestry
release of Aragam *et al.*[^aragam] (GWAS Catalog `GCST90132314`). **Every file is on GRCh38**: the
metabolite and CAD statistics are served already harmonised to GRCh38 by the GWAS Catalog, so no
coordinate liftover is applied to them.

The four exposures are chosen as a **gradient of causal-prior strength**, so the pipeline is
shown to *discriminate* rather than merely run. **LDL_C is the canonical positive control**: the
causal LDL→CAD relationship is established by the totality of genetic, epidemiologic, and
randomized-trial evidence — including RCTs of statins (*HMGCR*), PCSK9 inhibitors (*PCSK9*), and
ezetimibe (*NPC1L1*)[^ference] — so recovering a positive *cis*-MR at these drug-target genes is
the strongest validation of the whole arc. **BCAA** (anchored by the catabolic regulators *PPM1K* /
*BCKDK*), whose causal role in type 2 diabetes has itself been supported by Mendelian
randomization[^lotta], and **glucose** (anchored by *GCK* / *GCKR* / *G6PC2* / *SLC30A8*) are
further positive controls for the diabetes axis, both tested against T2D. **Lactate is the
contrast**: different biology (glycolysis), a weaker causal prior, and a notable locus at *GP6*
(platelet collagen receptor) biologically distant from a metabolic-disease mechanism — a test of
whether the methods correctly report a *less* compelling causal picture. Together the four exposures
span loci across the genome and several distinct pathways, so the pipeline can be judged on whether
it **discriminates** strong causal signals from weak ones.

[^tambets]: Tambets R, *et al.* Genome-wide association study for circulating metabolic traits in
619,372 individuals. *medRxiv* 2024.10.15.24315557 (2024); published in *Nature* (2026),
doi:10.1038/s41586-026-10532-5. Summary statistics via the NHGRI-EBI GWAS Catalog.

[^aragam]: Aragam KG, *et al.* Discovery and systematic characterization of risk variants and genes
for coronary artery disease in over a million participants. *Nat Genet* 54, 1803–1815 (2022),
doi:10.1038/s41588-022-01233-6. GWAS Catalog `GCST90132314` (European-ancestry release).

[^ference]: Ference BA, *et al.* Low-density lipoproteins cause atherosclerotic cardiovascular
disease. 1. Evidence from genetic, epidemiologic, and clinical studies. A consensus statement from
the European Atherosclerosis Society Consensus Panel. *Eur Heart J* 38(32), 2459–2472 (2017),
doi:10.1093/eurheartj/ehx144.

[^lotta]: Lotta LA, *et al.* Genetic predisposition to an impaired metabolism of the branched-chain
amino acids and risk of type 2 diabetes: a Mendelian randomisation analysis. *PLoS Med* 13(11),
e1002179 (2016), doi:10.1371/journal.pmed.1002179.

**Scope and resource constraints, stated up front.** Because of finite compute and a fixed
time budget, the analysis was **not** run across the full planned set of metabolites — the
report focuses on LDL_C (with the other three traits carried through the early steps) rather than
the paper's full metabolite panel. This is a scoping decision, not an architectural limit: the
pipeline is **config-driven** (`config/config.yaml`), so adding a metabolite is a one-line
accession entry and the same Snakemake rules then run end to end for it — the design scales to any
desired number of traits, and the restriction here is purely one of available resources. For the
same reason, some **tools and methodological choices deliberately diverge from the published
study** where a lighter or more self-contained alternative fit the resource envelope (these
substitutions are flagged in the affected sections). All work runs on a single workstation-class
machine (**8-core CPU, 8 GiB RAM, no GPU**); memory in particular drives several of these choices.

**The LD reference is out-of-sample.** Fine-mapping, clumping, and MR instrument pruning use a
1000 Genomes EUR panel, **not** the in-sample UKBB LD the reference paper used. Out-of-sample LD
can miscalibrate SuSiE credible sets and mis-prune instruments; this is an accepted tractability
tradeoff, flagged here and again in each affected section. With LDL_C selected as the metabolite of interest, **CAD is the
appropriate disease outcome** for the downstream colocalization and MR, so T2D is left out of the
reported arc.

# Meta-analysis (Step 1)

This step combines the two European cohorts — EstBB and UKBB_EUR — into a single genome-wide
association table for each metabolite using **inverse-variance-weighted, fixed-effect
meta-analysis**. The main steps are: each cohort's summary statistics are matched variant by
variant on chromosome, position, and both alleles (a **no-flip join**, so only identically
oriented effect alleles are combined and any strand or allele mismatch is left uncombined rather
than silently mis-merged); every variant is weighted by the inverse of its squared standard error,
so more precisely estimated cohorts contribute more; the pooled effect is the weighted mean of the
per-cohort betas, with a pooled standard error of √(1/Σw); and effect-allele frequency and sample
size are carried through, with a signed *z*-statistic, two-sided *p*-value, and a Cochran's-*Q*
heterogeneity test computed per variant. Crucially, **every union variant is kept** — the table is
not restricted to a HapMap3 or well-imputed subset — because the downstream steps (lead-variant
definition, fine-mapping, MR) need the full genome-wide signal. The whole meta is run
**one chromosome at a time and written incrementally**, so the full ~100M-variant table is produced
without ever holding more than a single chromosome in memory — the design choice that lets a
genome-wide meta run on an 8 GiB machine. Finally, the result is **validated against the authors'
published `meta_EUR`**: our pooled betas are regressed on theirs, and the OLS slope and Pearson *r*
should both sit at ≈1 if the allele harmonization and inverse-variance weighting are correct.

```{r meta-validation}
# Headline Step-1 result. Driven by the small precomputed `*.validation.tsv` (r, slope, variant
# overlap) — NOT by loading the multi-GB genome-wide meta files, which would exhaust memory.
val_tab <- read_many(sm@input$validation_tsv)
if (!is.null(val_tab) && nrow(val_tab)) {
  # tidy display: metabolite, comparison size, the two agreement metrics, cohort overlap
  keep <- intersect(c("metab","pearson_r","slope","n_variants","n_compared",
                      "n_both","n_estbb_only","n_ukbb_only"), names(val_tab))
  val_show <- val_tab[, ..keep]
  # keep full precision: the headline is that r/slope sit at machine precision to 1
  # (e.g. 0.999999999999989), so display ~15 sig figs rather than rounding to a bare "1".
  if ("pearson_r" %in% names(val_show)) val_show[, pearson_r := signif(pearson_r, 15)]
  if ("slope"     %in% names(val_show)) val_show[, slope     := signif(slope, 15)]
}
show_table(if (exists("val_show")) val_show else NULL,
           "Validation vs published meta_EUR: agreement (Pearson r, OLS slope) and cohort overlap")
```

The table reads out that validation. `pearson_r` and `slope` sit at machine precision to 1, meaning
the pooled effect sizes reproduce the authors' published meta exactly across all union variants,
while the overlap columns (`n_both`, `n_estbb_only`, `n_ukbb_only`) confirm the meta genuinely
combined both cohorts rather than defaulting to whichever one was present. Together these confirm
that the summary-statistic pre-processing was accurate and the genome-wide table is ready for the
downstream steps.

**How this compares to the source manuscript.** The reproduced European meta-analysis pools the
same two cohorts the authors describe — EstBB (n ≈ 185,352) and UKBB_EUR (n ≈ 413,897) — into a
`meta_EUR` of **n = 599,249**, matching the published sample size exactly. The validation scatter
here plays the same role as the manuscript's Supplementary Fig. 2, which regresses shared
lead-variant betas from their `meta_EUR` against a prior published analysis (Karjalainen et al.)
and reports near-perfect concordance where the samples overlap. The comparison differs in one
honest respect: the manuscript benchmarks against an *external* study with only partial sample
overlap, so their agreement is high but sub-unity; this replication benchmarks against the authors'
*own* published `meta_EUR` over the identical cohorts, so `r` and slope sit at machine precision to
1 — a reproduction check rather than a cross-study replication. Both confirm the same thing: allele
harmonisation and inverse-variance weighting recover effect sizes that agree with the reference.

```{r meta-fig, results='asis'}
# Two 2x2 composites (one image each) instead of 8 stacked plots.
# Validation (meta_rep vs meta_og) is the headline reproducibility check;
# concordance (EstBB vs UKBB per-cohort beta) is the supporting cross-cohort view.
cat("**Validation — reproduced meta (`meta_rep`) against the original published meta (`meta_og`), per metabolite:**\n\n")
compose_figs(sm@input$validation)
cat("**Cross-cohort concordance — EstBB vs UKBB effect sizes, per metabolite:**\n\n")
compose_figs(sm@input$concordance)
```

**Limitations.** This is a fixed-effect, inverse-variance-weighted meta-analysis, which assumes the
two cohorts estimate the same underlying effect. The per-variant Cochran's-*Q* is carried through so
heterogeneous loci can be flagged, but a fixed-effect pooled estimate can still be misleading where
the true effect genuinely differs between EstBB and UKBB. The no-flip join is deliberately
conservative — strand-ambiguous or allele-mismatched variants are left uncombined rather than risk a
silent mis-merge — so a small number of genuinely shared variants that a more aggressive
allele-flipping step could have harmonised are instead reported from a single cohort. Finally, the
validation here is a *reproduction* check against the authors' own published meta over the identical
cohorts: near-perfect agreement confirms the allele harmonisation and inverse-variance weighting are
correct, but it does not independently replicate the association signal in a fresh sample the way a
comparison against a separate study would.

## MCQ — Meta-analysis

**A variant is genome-wide significant in the EstBB + UKBB meta-analysis but reached significance
in neither cohort alone. The most appropriate interpretation is:**

- A. The meta-analysis has inflated type-I error, so the hit is probably false.
- B. Increased sample size raised power; concordant same-direction effects combined to cross the threshold.
- C. The variant must have opposite effects in the two cohorts.
- D. It indicates a batch effect between cohorts.

*Answer: **B.** Meta-analysis gains power precisely when per-cohort effects are concordant; a
sub-threshold-but-consistent signal crossing genome-wide significance is the expected behaviour,
not an artifact.*

# Heritability & genetic correlation (Step 2, LDSC)

With a genome-wide meta table in hand, this step characterises the *genetic architecture* of each
trait using LD Score Regression (LDSC), which estimates SNP-based heritability and pairwise genetic
correlation directly from summary statistics — no individual-level genotypes required, which is
exactly the constraint this pipeline operates under. Two products matter here. First,
per-(trait, dataset) SNP-heritability quantifies how much of each metabolite's variance is tagged by
common variants; computing it separately for EstBB, UKBB_EUR, and the meta gives a consistency check
across cohorts. Second, following the paper's design, two genetic-correlation analyses are run:
**Analysis A** correlates the three datasets (EstBB, UKBB_EUR, meta_EUR) for each trait, which tests
whether the two biobanks are measuring the same genetic signal and therefore whether meta-analysing
them is justified; and **Analysis B** correlates the four traits within each dataset, which places
the metabolites relative to one another.

The reasoning that shaped the step is that the estimates are only trustworthy if confounding is
controlled, so the **LDSC intercept** is read as the key diagnostic and reported alongside every h²
value. An intercept near 1 means the elevated test statistics reflect genuine polygenic signal
rather than population stratification or cryptic relatedness; a markedly elevated intercept would
instead flag confounding or sample overlap. This matters more than usual here because the pipeline
later relies on an out-of-sample LD reference, so establishing up front that inflation is polygenic
rather than artefactual is worth doing explicitly. On tooling, LDSC is run using the original `ldsc`
implementation (Python 2.7) in its own conda environment, with the summary statistics pre-restricted
to the ~1.2M HapMap3 rsIDs before munging — a memory-driven choice that keeps each job small enough
to run comfortably on the 8 GiB machine. This step runs across **all four traits**.

```{r ldsc-table}
# Read the PRECOMPUTED tables written by 02_ldsc_rg.R (single source of truth) rather than
# re-parsing the raw LDSC logs here — one parser, no drift between this report and Step 2.
h2_tab <- read_tsv1(sm@input$h2_tsv)
show_table(h2_tab, "LDSC SNP-heritability and intercept (per trait x dataset)")
rg_bb <- read_tsv1(sm@input$rg_bb)
show_table(rg_bb, "Analysis A — genetic correlation between biobanks, per trait")
rg_tr <- read_tsv1(sm@input$rg_tr)
show_table(rg_tr, "Analysis B — genetic correlation between traits, per dataset")
```

```{r ldsc-fig, results='asis'}
show_figs(c(sm@input$h2_plot, sm@input$rg_bb_plot, sm@input$rg_tr_plot))
```

The tables and figures read out as follows. Every LDSC intercept sits near 1, so the test-statistic
inflation is polygenic rather than confounded — the precondition for trusting the heritability
estimates. The four traits carry modest, plausible SNP-heritabilities (LDL_C highest, lactate
lowest), and Analysis A shows high between-biobank genetic correlation for each trait, which is the
empirical justification for having meta-analysed EstBB and UKBB in Step 1. Analysis B places the
traits relative to one another within each dataset. Together these confirm the meta table is
well-behaved and that the two cohorts share genetic architecture strongly enough to pool.

**How this compares to the source manuscript.** Two comparisons hold up against the paper's
genome-wide figures. *Heritability:* across their 249 metabolites the authors report SNP-h² ranging
from 2.8% (acetoacetate) to 19.5% (HDL size), median **10.2%** (their Supplementary Table 3 /
Supplementary Fig. 1). Our four traits fall inside that envelope — LDL_C at **11.75%** sits almost
exactly on their median, with BCAA (6.5%), Glucose (5.1%) and Lactate (3.9%) toward the
lower-heritability end, consistent with amino-acid and glycolysis metabolites being less heritable
than cholesterol traits in their data. *Cross-biobank genetic correlation:* the manuscript reports a
median r_g of **0.91** (mean 0.89) between EstBB and UKBB for matched metabolic traits, taking that
high shared genetic architecture as justification for meta-analysing the two biobanks. Our
between-biobank r_g (EstBB vs UKBB_EUR) is 0.85 (LDL_C), 0.83 (BCAA), 0.90 (Glucose) and 0.86
(Lactate) — mean **0.86**, the same "largely shared effects" conclusion on the four traits we ran.
The manuscript's estimate is a genome-wide median over 249 traits and ours a four-trait mean, so the
small gap is expected; the biological read-out is identical. The paper does not emphasise the LDSC
intercept the way this report does — that diagnostic is highlighted here because out-of-sample
issues (addressed at fine-mapping) make confounding-vs-polygenicity separation worth stating
explicitly.

**Limitations.** LDSC heritability is a *common-variant, HapMap3-tagged* estimate: pre-restricting to
the ~1.2M HapMap3 rsIDs is what makes the step tractable at 8 GiB, but it means rare-variant and
poorly-tagged contributions to heritability are not captured, so these h² values are lower bounds on
total SNP-heritability rather than complete accounts. LDSC also assumes a roughly polygenic
architecture and a well-matched LD reference; for traits with a few very large-effect loci (LDL_C
included) the regression can be sensitive to those loci, and the standard errors on a four-trait
panel are wider than the paper's genome-wide estimates over 249 traits. The genetic-correlation
estimates inherit the same caveats and, for Analysis A, the EstBB and UKBB_EUR inputs are not fully
independent of the meta they are correlated with, so those particular correlations are expected to
run high by construction.

## MCQ — Heritability (LDSC)

**The LDSC regression for a trait returns an intercept ≈ 1.0. This indicates:**

- A. Severe population stratification inflating all test statistics.
- B. Minimal confounding; inflation in the test statistics reflects genuine polygenic signal.
- C. The heritability estimate is invalid.
- D. Severe sample overlap between the two cohorts.

*Answer: **B.** The LDSC intercept isolates confounding/inflation from polygenicity; an intercept
near 1 means the elevated χ² comes from real polygenic signal, validating the h² estimate. (A
markedly elevated intercept would instead flag stratification or sample overlap.)*

# Lead variants (Step 3)

Having established that the meta table is well-behaved, this step reduces the genome-wide-significant
signal to a set of **independent lead variants** — one sentinel per locus — which become the
scaffold for fine-mapping. Two design choices define the step, and the reasoning behind each is worth
stating because both are deliberate departures from a naive threshold-and-clump.

The first is **frequency-dependent significance**. Common and low-frequency variants (MAF > 0.1%) are
called at the standard genome-wide line, P < 5×10⁻⁸, but rare variants (MAF < 0.1%) are held to a
stricter Bonferroni line, P < 0.05 / 8×10⁷ = 6.25×10⁻¹⁰. The reasoning is that the conventional
5×10⁻⁸ threshold is calibrated to roughly one million independent common-variant tests; the ~80M
rare-variant tests in a genome-wide metabolite scan make that line anti-conservative for rare
variants, so a tighter bar is needed to hold the family-wise error rate. Variants whose
effect-allele frequency is unknown fall back to the standard line and are flagged
`maf_class = "unknown"`, so a reader can weight them accordingly rather than having them silently
mixed in.

The second is a **distance-based locus definition**. The most significant unclaimed variant is taken
as a lead, a 2 Mb locus (±1 Mb) is claimed around it, every variant inside is removed, and the
process iterates until no genome-wide-significant variant remains. This is a considered substitute
rather than a shortcut: Step 3 operates on summary statistics only, with no genotype panel provided,
so pairwise r² between variants is not computable and true LD clumping is not available. The distance
rule is the standard, honest fallback in that situation. The paper's additional r²≥0.05 locus-merge —
which stops a single ultra-strong signal from "bleeding" past the 2 Mb window through long-range LD —
and its r²≥0.8 cross-trait clustering both require a EUR genotype panel, so they are deferred to the
LD-aware stage that Step 4 introduces.

```{r loci-table}
lead_tab <- read_many(sm@input$lead)
show_table(lead_tab, "Independent lead variants per metabolite")
```

```{r loci-fig, results='asis'}
show_figs(sm@input$manhattan)
```

The headline result of this step is the **concordance with the published loci**. Benchmarked against
the authors' `meta_EUR` LDL_C lead variants, this step recovers **96.3%** (310/322) of their loci at
the exact lead position — 321 leads called here against 322 published. That level of agreement, from
an independent reimplementation running on an out-of-sample LD reference, is the evidence that the
lead-calling is working. The small residual is exactly what an out-of-sample 1000G-EUR reference
predicts: within a shared association signal a different LD reference can tag a slightly different
sentinel SNP, so the locus is found but the precise lead variant occasionally differs by a few
kilobases. The Manhattan plot shows the genome-wide signal with the called leads marked.

**How this compares to the source manuscript.** Beyond the exact 322-vs-322 lead count already
shown, the frequency structure of the leads reproduces one of the paper's descriptive results. Their
Supplementary Fig. 3A plots lead-variant MAF against effect size across all `meta_EUR` locus–trait
pairs and shows the canonical inverse relationship — rare variants carry large effects, common
variants small ones — with point size scaled by −log₁₀P. The panel below is the LDL_C analogue built
from our own lead table: the same downward MAF–effect cline is present, the 14 rare leads
(MAF < 0.1%) occupy the high-effect corner, and the 307 common leads compress toward small effects.
PCSK9 (rs11591147, the R46L statin-target lead) is marked as the one fine-mapped coding sentinel.
This is a per-trait reproduction of a genome-wide manuscript figure, not the full 249-trait cloud,
so it is offered as a consistency check on the lead-calling rather than a like-for-like remake.

**Limitations.** The distance-only locus definition is the main scientific limitation of this step:
a single very strong signal (−log₁₀P ≫ 100) can carry association past the fixed ±1 Mb window through
LD, and a distance rule has no way to know two adjacent windows tag the same causal variant, so one
true signal can be split into several apparent "independent" loci. This is precisely what the paper's
r²≥0.05 merge exists to prevent, and it is only avoidable with a matched genotype panel — which Step 3
does not have. The frequency-dependent threshold, while defensible, uses a fixed ~80M denominator for
the rare-variant Bonferroni line rather than the true number of independent rare tests, so it is
approximate. And because effect-allele frequency is missing for some variants, the `unknown` class is
called at the looser line and may include a small number of rare variants that a complete
frequency annotation would have held to the stricter bar.

```{r loci-maf-effect-fig, results='asis'}
# LDL_C analogue of Karjalainen/paper Supplementary Fig. 3A: lead MAF vs |effect size|,
# point size = -log10 P, colour = frequency class. Built by 03_leadvars.R from the lead table.
show_figs(sm@input$maf_effect)
```

## MCQ — Lead variants

**Step 3 calls lead variants with a frequency-dependent threshold: P < 5×10⁻⁸ for common variants
(MAF > 0.1%) but the stricter P < 6.25×10⁻¹⁰ for rare variants. The rare-variant threshold is
tightened because:**

- A. Rare variants have larger effect sizes, so they need a higher bar.
- B. There are far more rare-variant tests (~80M), so the standard 5×10⁻⁸ line is anti-conservative and would admit too many false positives.
- C. Rare variants are always artifacts and should be suppressed.
- D. LDSC requires it for the heritability estimate.

*Answer: **B.** The genome-wide 5×10⁻⁸ threshold is calibrated to ~1M independent common-variant
tests. Adding ~80M rare-variant tests inflates the multiple-testing burden, so a Bonferroni-style
tightening (0.05/8×10⁷) is needed to hold the family-wise error rate.*

**A follow-on on locus definition: this replication defines independent loci by distance (top SNP,
then a ±1 Mb exclusion window, iterated) rather than by the paper's LD-based r²≥0.05 merge. The main
scientific limitation of the distance-only rule is:**

- A. It cannot handle more than one chromosome.
- B. A single very strong signal (−log₁₀P ≫ 100) can generate association that extends beyond 2 Mb through LD, so distance clumping may split one true signal into several "independent" loci.
- C. It systematically undercounts loci relative to LD clumping.
- D. It requires an in-sample LD matrix to run.

*Answer: **B.** LD can carry a strong association past a fixed window; the paper's r²≥0.05 merge
exists specifically to stop that "bleeding" and collapse the fragments into one locus. Distance
clumping has no way to know two windows tag the same causal variant, so it can over-count. It is the
honest choice only when no matched genotype panel is available to compute pairwise r² — which is the
case here.*

# Fine-mapping + coding annotation (Step 4)

The lead variants mark *where* the associations are; this step asks *which variant* at each locus is
most plausibly causal, and *what gene* it acts through. Each locus is fine-mapped with **SuSiE**
(the summary-statistics form, SuSiE-RSS), which decomposes the regional signal into one or more
credible sets, each a small set of variants with a posterior inclusion probability (PIP) that one of
them is causal. The credible-set variants are then annotated for coding consequence (missense via
Ensembl VEP), so a fine-mapped signal can be connected to a candidate effector gene. The MHC region
is excluded because its extreme, long-range LD breaks the fine-mapping assumptions.

The scope here is a deliberate, reasoned narrowing. Rather than fine-map genome-wide, this step
targets the **11 canonical LDL effector loci for LDL_C only**. The thinking is that LDL_C is the
positive control and these 11 loci have known biology, so restricting to them lets the out-of-sample
LD behaviour be inspected locus by locus — where a credible set is well-behaved and where it is not
— instead of being averaged into a genome-wide count. Two SuSiE settings follow directly from the
out-of-sample-LD concern: `estimate_residual_variance` is fixed to `FALSE`, and an `estimate_s_rss`
LD-mismatch diagnostic is computed per locus, both of which are the documented remedies for running
SuSiE-RSS against a reference panel that is not the GWAS sample.

```{r finemap-table}
annot_tab <- read_many(sm@input$annot)
show_table(annot_tab, "Credible-set variants and coding annotation")
```

Only the credible-set variants annotated as **missense** are shown below — these are the coding
candidates that fine-mapping nominates as potentially causal. The recovery of textbook LDL coding
variants (PCSK9 R46L, APOE C130R) at PIP = 1 is the strongest validation that the pipeline works
despite the out-of-sample LD.

**How this compares to the source manuscript.** The authors fine-map genome-wide across all traits
and report 116,467 credible sets, of which 31,392 (27%) resolve to 3,000 distinct putative causal
variants at PIP > 0.8; these include 271 missense variants (Ensembl VEP) and 172 predicted
splice-altering variants (their Supplementary Table 4). This replication is deliberately narrower:
it fine-maps the 11 canonical LDL effector loci for LDL_C only — a hypothesis-driven subset chosen
so out-of-sample-LD behaviour can be inspected locus by locus — yielding 68 credible-set variants at
PIP > 0.8, of which 5 are missense (PCSK9 R46L, ABCG5 R198Q, ABCG8 T400K, APOB T98I, APOE C130R).
The honest comparison is therefore one of *class membership*, not counts: our five coding candidates
belong to the same missense class the manuscript catalogues genome-wide, and each is an established
lipid effector variant, so recovering them at high PIP shows the fine-mapping mechanism is calibrated
even though the scope is a fraction of the paper's. The manuscript's showcase example — a fine-mapped
missense variant in BCAT2 (19-48806519-G-C, PIP = 1) that is also splice-altering and drives BCAA
levels — sits outside our LDL_C scope and is not something this run reproduces; it is noted as
context for the kind of coding+splicing dual mechanism the full analysis surfaces. Splice annotation
(SpliceAI / AlphaGenome) is not part of this replication, so the paper's 172 splice-altering variants
have no analogue here.

```{r finemap-missense}
# Highlight the coding candidates. annot.tsv carries every credible-set variant; the missense
# subset is the biologically interpretable headline.
if (!is.null(annot_tab) && nrow(annot_tab) && "is_missense" %in% names(annot_tab)) {
  keep <- intersect(c("gene_symbol","ID","aa_change","pip","cs","sift_pred",
                      "polyphen_pred","is_known_causal"), names(annot_tab))
  miss <- annot_tab[is_missense == TRUE, ..keep][order(-pip)]
  show_table(miss, "Missense credible-set variants (coding effector candidates)")
} else {
  show_table(NULL)
}
```

```{r finemap-fig, results='asis'}
# Two cross-locus summary figures (per-locus QC counts; coding-vs-regulatory consequence mix).
show_figs(c(sm@input$finemap_summary, sm@input$consequence_summary), cols = 1)
```

```{r finemap-regional-fig, results='asis'}
# LocusZoom-style regional plots (GWAS -log10P coloured by r2-to-lead over a PIP track).
# We FEATURE three loci that each make a distinct point rather than showing all 11 — the
# two summary figures above already give the all-loci view, and 11 near-identical panels
# add scroll without insight:
#   PCSK9   - the R46L drug-target missense variant recovered at high PIP (headline result)
#   ABCG5_8 - two fine-mapped missense variants across a two-gene locus (coding + naming nuance)
#   APOE    - the canonical lipid locus with multiple credible sets (allelic heterogeneity)
featured <- c("PCSK9", "ABCG5_8", "APOE")
plotdir  <- sm@input$finemap_plots[1]
regional_pngs <- file.path(plotdir, sprintf("%s.%s.regional.png", "LDL_C", featured))
cat("Three representative loci are featured here, of the 11 fine-mapped. PCSK9 shows the R46L",
    "drug-target coding variant recovered at high PIP; ABCG5/ABCG8 shows two missense variants",
    "across a two-gene locus; and APOE shows a locus resolved into multiple credible sets",
    "(allelic heterogeneity). The remaining loci are summarised in the two figures above and",
    "available in the repository.\n\n")
show_figs(regional_pngs, cols = 1)
```

**Limitations.** The out-of-sample 1000G EUR LD reference is the single most important caveat of the
whole analysis, and it bites hardest here: SuSiE-RSS assumes the LD matrix matches the GWAS sample,
and a mismatched reference can miscalibrate PIPs and produce spurious or oversized credible sets. The
fingerprint of that mismatch is visible in the convergence QC — most loci hit the IBSS iteration cap
— which is why the LD-mismatch diagnostic is reported per locus and the residual variance is fixed
rather than estimated. Two further limits bound what this step can claim. The scope is 11 LDL effector
loci for one trait, so it is a hypothesis-driven probe of the method's calibration, not a
genome-wide fine-mapping result. And coding annotation explains only a minority of the signal: of the
credible-set variants only a handful are missense, while most are regulatory or intronic and cannot
be assigned an effector gene from sequence consequence alone — resolving those requires the
molecular-QTL colocalization of the next step. Splice-effect prediction (SpliceAI / AlphaGenome) is
not part of this replication, so splice-altering credible-set variants are not identified here.

## MCQ — Fine-mapping

**The pipeline fine-maps with an out-of-sample 1000G EUR LD reference rather than the in-sample
UKBB LD used by the reference paper. The primary risk this introduces to the credible sets is:**

- A. They will always be too small.
- B. LD mismatch can miscalibrate posterior inclusion probabilities and yield spurious or oversized credible sets.
- C. SuSiE cannot run at all without in-sample LD.
- D. It only affects the Manhattan plot, not fine-mapping.

*Answer: **B.** SuSiE-RSS assumes the LD matrix matches the GWAS sample; a mismatched reference
distorts PIPs and credible-set membership. In this run the fingerprint is visible in the
convergence QC (most loci hit the IBSS iteration cap), which is why an `estimate_s_rss` LD-mismatch
diagnostic is reported per locus and `estimate_residual_variance` is fixed to `FALSE` — the
documented remedy for out-of-sample LD. This is the single most important caveat of the analysis.*

# Colocalization with molecular QTLs

Fine-mapping (Step 4) narrows each locus to a short list of candidate causal variants, but a
credible set does not by itself name the gene through which a variant acts. Colocalization answers
that question: it asks whether the LDL-C association and a molecular-QTL association at the same
locus are driven by *one shared causal variant* — evidence that the trait signal acts through that
molecular phenotype (a gene's expression, splicing, or protein level) — or by two *distinct* variants
that happen to sit nearby. This is the step that converts a statistical locus into a candidate
effector gene with a mechanism attached.

**Analysis and scope.** For each of the LDL-C effector loci I tested the GWAS signal against
molecular QTLs using `coloc.abf` (the single-causal-variant approximate Bayes factor model) with the
same priors the reference paper adopted (`p1 = p2 = 1e-4`, `p12 = 5e-6`). Two design choices carry
the reasoning of this step. First, **tissue is chosen from the data, not assumed.** Liver is the
biologically obvious tissue for LDL, but the eQTL Catalogue has only one liver dataset (GTEx, n=208)
and it fine-maps none of these genes — so, exactly as the reference paper (a genome-wide `gpu-coloc`
scan over the whole Catalogue) and Open Targets do, I scan every gene-level dataset across all seven
quantification methods (`ge`, `exon`, `tx`, `txrev`, `leafcutter`, `microarray`, and `aptamer`
protein levels) and colocalize each gene against the best-powered dataset that actually fine-maps it.
Second, **the molecular layer is broad on purpose.** Restricting to bulk gene-expression QTLs would
miss the mechanism at loci that act through splicing (captured by `exon`/`leafcutter`) or through
circulating protein level (captured by the `aptamer` pQTLs) — so all quant methods are in scope and
the winning method is reported per gene. The scope of this replication is deliberately narrower than
the paper in one respect: the published study embeds colocalization in a disease → trait →
molecular-QTL triplet graph, whereas this pipeline has no disease (e.g. coronary artery disease)
layer, so I recover the trait ↔ molecular-QTL half of that graph at the canonical LDL effector loci.

```{r coloc-table}
coloc_best <- read_many(sm@input$coloc_best)
if (!is.null(coloc_best) && nrow(coloc_best)) {
  pp4 <- grep("PP.H4", names(coloc_best), fixed = TRUE, value = TRUE)[1]
  pp3 <- grep("PP.H3", names(coloc_best), fixed = TRUE, value = TRUE)[1]
  pp1 <- grep("PP.H1", names(coloc_best), fixed = TRUE, value = TRUE)[1]
  coloc_best[, Outcome := fifelse(get(pp4) >= 0.80, "colocalizes (shared causal variant)",
                          fifelse(get(pp3) >= 0.50, "distinct causal variants",
                          fifelse(get(pp1) >= 0.50, "GWAS-only (QTL null in tissue)", "ambiguous")))]
  tab <- coloc_best[order(-get(pp4)),
    .(Gene = symbol, `QTL type` = quant, Tissue = qtl,
      `PP.H4` = round(get(pp4), 3), `PP.H3` = round(get(pp3), 3),
      `shared SNPs` = nshared, Outcome)]
  show_table(tab, "Best-quant colocalization per LDL-C effector gene (PP.H4 = shared causal variant)")
}
```

**Results.** Of the 12 effector genes carried forward (each tested against its best-powered
fine-mapping dataset, out of 45 gene × quantification-method candidates in total), **five colocalize
at PP.H4 ≥ 0.80**: HMGCR (PP.H4 = 1.00, skin exon-QTL), APOA5 (0.99, plasma protein level),
CETP (0.99, lung expression), ABCG5 (0.97, spleen expression) and PCSK9 (0.94, tibial-nerve
expression). The two textbook druggable LDL targets in the set — HMGCR (the statin target) and
PCSK9 (the target of evolocumab/alirocumab) — both colocalize, which is the positive control that
tells us the pipeline is finding real shared signals rather than noise. The remaining genes split
into two informative classes. Four show a high probability of **distinct causal variants** (PP.H3):
ABCA1 (0.99), APOB (0.96), LPL (0.85) and APOE (0.72) — the trait and the molecular QTL both
associate at the locus, but through different causal variants, so the tested molecular phenotype in
that tissue is not the mechanism. APOB is the instructive case: it carries strong coding-haplotype
biology, and the LDL signal there is not explained by the skin exon-QTL, consistent with a direct
protein-coding rather than an expression-regulatory mechanism. The last three — ABCG8, LDLR and
NPC1L1 — are **GWAS-only** (high PP.H1): the molecular QTL is simply not associated in the tissue
tested, most likely a power or tissue-context limitation rather than evidence against the gene.

```{r coloc-overview-fig, results='asis'}
cat("The colocalization outcome for every effector gene, decomposing the posterior into PP.H4",
    "(one shared causal variant, red) versus PP.H3 (two distinct causal variants, grey). The",
    "dashed line marks the PP.H4 = 0.80 colocalization threshold. Genes separate cleanly into the",
    "five that colocalize on a shared variant and the loci that resolve to distinct signals.\n\n")
show_figs(sm@input$coloc_overview)
```

```{r coloc-regional-fig, results='asis'}
cat("Regional colocalization plots for the five colocalizing genes, each in the reference paper's",
    "Supplementary-Figure-7 style: the LDL-C GWAS association (top) stacked over the molecular-QTL",
    "association (bottom) on a shared genomic axis, with the lead variant of each panel marked and",
    "PP.H4 annotated. A shared causal variant shows up as the two peaks coinciding. This is a",
    "two-panel analogue of the paper's three-panel figure — we have no disease-GWAS layer to add",
    "on top.\n\n")
coloc_regional <- list.files(sm@input$coloc_plots, pattern = "\\.regional\\.png$", full.names = TRUE)
show_figs(coloc_regional, cols = 2)
```

**How this compares to the source manuscript.** The correspondence is at the level of method and
priors, not counts. The paper runs `gpu-coloc` genome-wide over SuSiE-fine-mapped Bayes factors
(`coloc.bf_bf`) with priors `p1 = p2 = 1e-4`, `p12 = 5e-6`; this replication uses the single-signal
`coloc.abf` on marginal statistics with the identical priors, which is the deliberately simpler,
CPU-tractable case of the same model. Because the published analysis is genome-wide and
disease-anchored while this one is a targeted trait-versus-molecular-QTL sweep over a handful of LDL
effector loci, the honest comparison is *shared-signal recovery at known loci*, and it holds: the
flagship recreation is HMGCR, where the LDL-C GWAS lead **rs12916** and the skin exon-QTL peak both
resolve to the same narrow region — the QTL peak sits on **5-75355259**, the very variant the paper's
Supplementary Figure 7 highlights as the fine-mapped causal SNP (PIP ≈ 0.64) — reproducing that
figure's visual signature of a PP.H4 ≈ 1.0 colocalization from an independent implementation.

**Limitations.** Colocalization is only as informative as the molecular data behind it, and three
limits bound what this step can claim. First, **it is tissue- and context-specific**: a null result
(the three GWAS-only genes) means the QTL was not associated in the *tissue tested*, not that the gene
is not causal — the true effector tissue may simply be absent or underpowered in the Catalogue, which
is precisely why liver could not be used for any of these genes. Second, **`coloc.abf` assumes a
single causal variant per locus**; at loci with allelic heterogeneity (which the Step-4 fine-mapping
already showed exists here) that assumption can understate a true colocalization or misassign the
shared variant, and the paper's SuSiE-based `coloc.bf_bf` is the more faithful model for exactly that
reason. Third, **the GWAS side inherits the out-of-sample-LD caveat** from the whole analysis — the
marginal statistics are unaffected, but any locus where fine-mapping was already shaky is one to read
with corresponding caution. Finally, a shared causal variant is strong evidence of a mechanism but not
proof of causality for the trait; that step is Mendelian randomization (Steps 6–7).

## MCQ — Colocalization

**At the ABCG8 / LDLR / NPC1L1 loci the LDL-C GWAS signal is genome-wide significant, yet the
colocalization posterior puts most of its mass on PP.H1 (GWAS-only), not PP.H4 or PP.H3. What is the
correct interpretation?**

- A. These genes are not causal for LDL cholesterol.
- B. The colocalization failed and the run should be discarded.
- C. The molecular QTL is not associated in the tissue/assay tested, so there is no QTL signal to colocalize with — a power/tissue-context limitation, not evidence against the gene.
- D. PP.H1 means the two signals share a causal variant.

*Answer: **C.** PP.H1 is the posterior for "association in the GWAS only, none in the QTL dataset."
It arises when the molecular QTL simply is not detected in the tissue or assay available — a
limitation of QTL power and tissue coverage (recall no liver dataset fine-maps any of these genes),
not a statement that the gene is not causal. LDLR is the definitive counter-example: it is the
canonical LDL-clearance receptor and unquestionably causal, yet it does not colocalize here because
its expression QTL is not associated in the tested tissue. This is why a colocalization null must be
read as "no evidence in this molecular layer," never as "the gene is not causal" (A), a failed run
(B), or a shared signal (D, which is PP.H4).*

# Limitations

- **Out-of-sample LD.** Fine-mapping and distance-clumping use a 1000G EUR reference rather than
  in-sample LD; SuSiE credible sets may be miscalibrated (flagged per locus by the kriging
  diagnostic). This is the single most important caveat and the main methodological divergence
  from the reference paper, which had in-sample UKBB LD.
- **Distance-only locus definition.** With no genotype panel provided at Step 3, independent loci
  are defined by a ±1 Mb distance rule rather than the paper's r²≥0.05 LD merge; a very strong
  signal can bleed past the window and be split into apparent multiple loci (see the Step-3 MCQ).
- **Fine-mapping scoped to LDL effector loci.** Step 4 fine-maps the 11 canonical LDL effector
  loci (a hypothesis-driven subset), not genome-wide — chosen so out-of-sample-LD behaviour can be
  inspected locus by locus.
- **Coding annotation explains few loci.** Of the credible-set variants, only a handful are
  missense; most signal is regulatory/intronic. Resolving those mechanisms needs molecular-QTL
  colocalization (Step 5), which recovers a shared causal variant at five effector genes.
- **Colocalization is tissue- and model-limited.** `coloc.abf` assumes one causal variant per locus,
  and a null result reflects the tissue/assay tested rather than evidence against the gene; the
  reference paper's SuSiE-based `coloc.bf_bf` over the full eQTL Catalogue is the more complete model.
- **EUR-only.** Exposures and (planned) outcomes are European-ancestry, so estimates may not
  generalize — a deliberate choice to keep LD and allele frequencies matched.

*Mendelian randomization (Steps 6–7) is in progress and will be added to this report, with its own
MCQ, once its pipeline rules have run.*

