---
title: "VariantBench — Metabolic-Trait GWAS Follow-up (LDL_C, BCAA, Glucose, Lactate; EUR)"
author: "Khushi Goda"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
library(data.table); library(ggplot2); library(DT)
library(cowplot); library(png)   # compose_figs(): tile per-trait PNGs into one image
sm <- snakemake   # S4 object; inputs wired by the `report` rule in the Snakefile
set.seed(sm@params$seed)   # reproducibility: global seed from config

# ---- helpers (make tomorrow "plug data in and knit") ----------------------
# Read one tsv defensively: empty/absent -> NULL (so scaffold runs never error).
read_tsv1 <- function(path) tryCatch(fread(path), error = function(e) NULL)

# Read several tsvs and stack them, tagging each row with its source file.
read_many <- function(paths) {
  rbindlist(lapply(paths, function(p) {
    d <- read_tsv1(p)
    if (!is.null(d) && nrow(d)) d[, source_file := basename(p)]
    d
  }), fill = TRUE, use.names = TRUE)
}

# Render a table, or a placeholder line if there are no rows yet.
show_table <- function(dt, caption = NULL) {
  if (is.null(dt) || nrow(dt) == 0) {
    cat("*No rows yet — this table populates when the pipeline runs on real data.*\n\n")
    return(invisible(NULL))
  }
  DT::datatable(dt, caption = caption, rownames = FALSE,
                options = list(pageLength = 10, scrollX = TRUE))
}

# Embed figures that actually exist (skips scaffold-empty paths cleanly).
# NB: we DON'T use include_graphics here. Snakemake renders the Rmd from a temp
# script dir (.snakemake/scripts/), and both include_graphics and Pandoc's
# self_contained pass resolve image paths relative to THAT dir — so any path we pass
# (relative -> "Could not fetch resource"; absolute -> relativised to "../../results/..")
# fails. Instead we read each file (cwd = knit_root_dir = repo root, so file.exists
# passes) and inline it as a base64 data-URI <img>, which needs no path resolution at
# all. Chunks calling this MUST set results='asis' so the HTML is emitted raw.
show_figs <- function(paths, cols = NULL) {
  paths <- paths[file.exists(paths) & file.info(paths)$size > 0]
  if (!length(paths)) return(invisible(NULL))
  # A single figure renders full-width & centred; multiple figures render in a
  # responsive CSS grid (default 2 cols, auto-fit down to 1 on narrow screens) so
  # groups like the 11 regional LocusZoom panels read as a compact gallery instead
  # of an endless vertical scroll.
  if (is.null(cols)) cols <- if (length(paths) > 1) 2 else 1
  if (cols <= 1 || length(paths) == 1) {
    for (p in paths)
      cat(sprintf('<img src="%s" style="max-width:100%%;display:block;margin:0.5em auto;" alt="%s"/>\n\n',
                  knitr::image_uri(p), basename(p)))
  } else {
    cat(sprintf('<div style="display:grid;grid-template-columns:repeat(%d,1fr);gap:0.6em;margin:0.5em 0;">\n', cols))
    for (p in paths)
      cat(sprintf('<img src="%s" style="width:100%%;height:auto;" alt="%s"/>\n',
                  knitr::image_uri(p), basename(p)))
    cat('</div>\n\n')
  }
  invisible(NULL)
}

# Tile several PNGs into ONE composite image (cowplot) and inline it base64.
# Used for the Step-1 per-trait scatter families (4 validation, 4 concordance):
# one 2x2 image reads far better than 8 separate full-width plots and cuts weight.
# Each source PNG already carries its own trait title, so no cowplot labels.
compose_figs <- function(paths, ncol = 2, panel_in = 4.5) {
  paths <- paths[file.exists(paths) & file.info(paths)$size > 0]
  if (!length(paths)) return(invisible(NULL))
  if (length(paths) == 1) return(show_figs(paths))
  grobs <- lapply(paths, function(p)
    ggdraw() + draw_grob(rasterGrob(readPNG(p), interpolate = TRUE)))
  g <- plot_grid(plotlist = grobs, ncol = ncol)
  nrow <- ceiling(length(paths) / ncol)
  tmp <- tempfile(fileext = ".png")
  ggsave(tmp, g, width = panel_in * ncol, height = panel_in * nrow,
         dpi = 150, device = grDevices::png, type = "cairo", bg = "white")
  cat(sprintf('<img src="%s" style="max-width:100%%;display:block;margin:0.5em auto;" alt="composite"/>\n\n',
              knitr::image_uri(tmp)))
  invisible(NULL)
}
```

# Overview & method choices

> **Reproducibility.** This document is the R Markdown source `workflow/scripts/08_report.Rmd`,
> knitted by the Snakemake `report` rule. Every number, table, and figure below is generated
> at knit time by reading the pipeline's output files in `results/` — nothing is hand-entered
> or pasted. Re-running the pipeline and re-knitting regenerates this report end to end, so the
> report cannot drift from the analysis it describes.

> **Scope of this knit.** This version reports the **completed steps: meta-analysis, LDSC,
> lead-variant definition, and fine-mapping + coding annotation (Steps 1–4)**, each followed by a
> multiple-choice question testing the biological reasoning behind that step. Steps 1–2 run across
> all four traits; Steps 3–4 are shown for **LDL_C**, the positive-control trait whose 11 canonical
> effector loci let the out-of-sample-LD behaviour be inspected locus by locus. Colocalization and
> Mendelian randomization (Steps 5–7) are in progress and will be folded into this same report —
> each with its own MCQ — once their pipeline rules have run.

This report follows four circulating metabolic traits — **LDL cholesterol (LDL_C)**, **total
branched-chain amino acids (BCAA)**, **glucose**, and **lactate** — from published GWAS summary
statistics through a standard post-GWAS causal-inference chain: cross-cohort meta-analysis,
heritability and genetic correlation (LDSC), lead-variant definition, statistical fine-mapping,
colocalization with gene expression, and Mendelian randomization against two cardiometabolic
diseases. The analysis is **European-ancestry throughout** (EstBB + UKBB_EUR exposures,
EUR-matched CAD/T2D outcomes, 1000G EUR LD reference), and **no individual-level genotype data
is used** — every step operates on summary statistics plus a public LD reference.

The four exposures are chosen as a **gradient of causal-prior strength**, so the pipeline is
shown to *discriminate* rather than merely run. **LDL_C is the canonical positive control**: the
causal LDL→CAD relationship is proven by RCTs of statins (*HMGCR*), PCSK9 inhibitors (*PCSK9*),
and ezetimibe (*NPC1L1*), so recovering a positive *cis*-MR at these drug-target genes is the
strongest validation of the whole arc. **BCAA** (anchored by the catabolic regulators *PPM1K* /
*BCKDK*) and **glucose** (anchored by *GCK* / *GCKR* / *G6PC2* / *SLC30A8*) are further positive
controls for the diabetes axis, both tested against T2D. **Lactate is the contrast**: different
biology (glycolysis), a weaker causal prior, and a notable locus at *GP6* (platelet collagen
receptor) biologically distant from a metabolic-disease mechanism — a test of whether the methods
correctly report a *less* compelling causal picture. Four exposures across four chromosomes and
pathways demonstrate that the pipeline **discriminates** strong causal signals from weak ones.

**Method choice stated up front — the LD reference is out-of-sample.** Fine-mapping,
clumping, and MR instrument pruning use a 1000 Genomes EUR panel, **not** the in-sample UKBB
LD the reference paper used. Out-of-sample LD can miscalibrate SuSiE credible sets and
mis-prune instruments; this is an accepted tractability tradeoff, flagged here and again in
each affected section. A second deliberate choice concerns the CAD outcome: the paper's
Methods cite the multi-ancestry `GCST90132315`, but the authors' released code uses the
European-only `GCST90132314`. I use the European-only accession — it keeps exposure and
outcome ancestry-matched and avoids a known source of two-sample MR bias, and it is the
choice the authors' own code made. (The multi-ancestry study is not used: with EUR-only
exposures it would reintroduce the ancestry mismatch this design is built to avoid.) The
CAD file is the GWAS Catalog harmonised (GRCh38) release, matching the GRCh38 metabolite
exposures, so no CAD liftover is required; T2D (DIAGRAM, GRCh37) is lifted to GRCh38 in Step 0.

# Meta-analysis (Step 1)

*Inverse-variance fixed-effect meta-analysis of EstBB + UKBB (EUR), genome-wide (every union
variant kept, not HapMap3-restricted). Validation: my meta effect sizes are regressed against
the authors' published meta-analysis — the OLS slope and Pearson r should both be ≈1 if the
allele harmonization and inverse-variance weighting are correct.*

```{r meta-validation}
# Headline Step-1 result. Driven by the small precomputed `*.validation.tsv` (r, slope, variant
# overlap) — NOT by loading the multi-GB genome-wide meta files, which would exhaust memory.
val_tab <- read_many(sm@input$validation_tsv)
if (!is.null(val_tab) && nrow(val_tab)) {
  # tidy display: metabolite, comparison size, the two agreement metrics, cohort overlap
  keep <- intersect(c("metab","pearson_r","slope","n_variants","n_compared",
                      "n_both","n_estbb_only","n_ukbb_only"), names(val_tab))
  val_show <- val_tab[, ..keep]
  # keep full precision: the headline is that r/slope sit at machine precision to 1
  # (e.g. 0.999999999999989), so display ~15 sig figs rather than rounding to a bare "1".
  if ("pearson_r" %in% names(val_show)) val_show[, pearson_r := signif(pearson_r, 15)]
  if ("slope"     %in% names(val_show)) val_show[, slope     := signif(slope, 15)]
}
show_table(if (exists("val_show")) val_show else NULL,
           "Validation vs published meta_EUR: agreement (Pearson r, OLS slope) and cohort overlap")
```

The validation table is the decisive check on the whole meta step: `pearson_r` and `slope` at
machine precision to 1 mean the from-scratch inverse-variance meta reproduces the authors'
published meta exactly, over all union variants. The overlap columns (`n_both`,
`n_estbb_only`, `n_ukbb_only`) show that the meta genuinely combined both cohorts rather than
defaulting to whichever one was present. *(Counting genome-wide-significant loci is deferred to
Step 3, where LD-clumping collapses correlated hits into independent lead variants — a raw
p < 5e-8 variant count over 100M+ variants over-counts single signals through LD.)*

```{r meta-fig, results='asis'}
# Two 2x2 composites (one image each) instead of 8 stacked plots.
# Validation (our beta vs published meta_EUR) is the headline reproducibility check;
# concordance (EstBB vs UKBB per-cohort beta) is the supporting cross-cohort view.
cat("**Validation — our meta vs the authors' published `meta_EUR` (per trait):**\n\n")
compose_figs(sm@input$validation)
cat("**Cross-cohort concordance — EstBB vs UKBB effect sizes (per trait):**\n\n")
compose_figs(sm@input$concordance)
```

## MCQ — Meta-analysis

**A variant is genome-wide significant in the EstBB + UKBB meta-analysis but reached significance
in neither cohort alone. The most appropriate interpretation is:**

- A. The meta-analysis has inflated type-I error, so the hit is probably false.
- B. Increased sample size raised power; concordant same-direction effects combined to cross the threshold.
- C. The variant must have opposite effects in the two cohorts.
- D. It indicates a batch effect between cohorts.

*Answer: **B.** Meta-analysis gains power precisely when per-cohort effects are concordant; a
sub-threshold-but-consistent signal crossing genome-wide significance is the expected behaviour,
not an artifact.*

# Heritability & genetic correlation (Step 2, LDSC)

*SNP-heritability per (trait, dataset) and the paper's two genetic-correlation analyses —
**A**: rg between the three datasets (EstBB, UKBB_EUR, meta_EUR) for each trait; **B**: rg
between the four traits within each dataset. The **LDSC intercept** is the key diagnostic:
an intercept ≈1 means test-statistic inflation reflects polygenic signal, not population
stratification or cryptic relatedness.*

```{r ldsc-table}
# Read the PRECOMPUTED tables written by 02_ldsc_rg.R (single source of truth) rather than
# re-parsing the raw LDSC logs here — one parser, no drift between this report and Step 2.
h2_tab <- read_tsv1(sm@input$h2_tsv)
show_table(h2_tab, "LDSC SNP-heritability and intercept (per trait x dataset)")
rg_bb <- read_tsv1(sm@input$rg_bb)
show_table(rg_bb, "Analysis A — genetic correlation between biobanks, per trait")
rg_tr <- read_tsv1(sm@input$rg_tr)
show_table(rg_tr, "Analysis B — genetic correlation between traits, per dataset")
```

```{r ldsc-fig, results='asis'}
show_figs(c(sm@input$h2_plot, sm@input$rg_bb_plot, sm@input$rg_tr_plot))
```

## MCQ — Heritability (LDSC)

**The LDSC regression for a trait returns an intercept ≈ 1.0. This indicates:**

- A. Severe population stratification inflating all test statistics.
- B. Minimal confounding; inflation in the test statistics reflects genuine polygenic signal.
- C. The heritability estimate is invalid.
- D. Severe sample overlap between the two cohorts.

*Answer: **B.** The LDSC intercept isolates confounding/inflation from polygenicity; an intercept
near 1 means the elevated χ² comes from real polygenic signal, validating the h² estimate. (A
markedly elevated intercept would instead flag stratification or sample overlap.)*

# Lead variants (Step 3)

*Genome-wide-significant loci reduced to independent lead variants. Two design choices define
this step:*

*1. **Frequency-dependent significance.** Common/low-frequency variants (MAF > 0.1%) use the
standard genome-wide line P < 5e-8; rare variants (MAF < 0.1%) use a stricter Bonferroni line
P < 0.05/8e7 = 6.25e-10, because the ~80M rare tests make 5e-8 anti-conservative. Variants whose
effect-allele frequency is unknown fall back to the standard line and are flagged `maf_class =
"unknown"` so the reader can weight them accordingly.*

*2. **Distance-based locus definition.** The most-significant unclaimed variant is taken as a
lead, a 2 Mb locus (±1 Mb) is claimed around it, all variants inside are removed, and the process
iterates. The brief's Step 3 is summary-statistics-only and provides no genotype panel, so pairwise
r² is not computable; the distance rule is the standard, honest substitute. The paper's additional
r²≥0.05 locus-merge (which stops an ultra-strong signal from "bleeding" past the 2 Mb window) and
r²≥0.8 cross-trait clustering require a EUR genotype panel and are deferred to the LD-aware
follow-on (Step 4 sourcing).*

```{r loci-table}
lead_tab <- read_many(sm@input$lead)
show_table(lead_tab, "Independent lead variants per metabolite")
```

```{r loci-fig, results='asis'}
show_figs(sm@input$manhattan)
```

## MCQ — Lead variants

**Step 3 calls lead variants with a frequency-dependent threshold: P < 5×10⁻⁸ for common variants
(MAF > 0.1%) but the stricter P < 6.25×10⁻¹⁰ for rare variants. The rare-variant threshold is
tightened because:**

- A. Rare variants have larger effect sizes, so they need a higher bar.
- B. There are far more rare-variant tests (~80M), so the standard 5×10⁻⁸ line is anti-conservative and would admit too many false positives.
- C. Rare variants are always artifacts and should be suppressed.
- D. LDSC requires it for the heritability estimate.

*Answer: **B.** The genome-wide 5×10⁻⁸ threshold is calibrated to ~1M independent common-variant
tests. Adding ~80M rare-variant tests inflates the multiple-testing burden, so a Bonferroni-style
tightening (0.05/8×10⁷) is needed to hold the family-wise error rate.*

**A follow-on on locus definition: this replication defines independent loci by distance (top SNP,
then a ±1 Mb exclusion window, iterated) rather than by the paper's LD-based r²≥0.05 merge. The main
scientific limitation of the distance-only rule is:**

- A. It cannot handle more than one chromosome.
- B. A single very strong signal (−log₁₀P ≫ 100) can generate association that extends beyond 2 Mb through LD, so distance clumping may split one true signal into several "independent" loci.
- C. It systematically undercounts loci relative to LD clumping.
- D. It requires an in-sample LD matrix to run.

*Answer: **B.** LD can carry a strong association past a fixed window; the paper's r²≥0.05 merge
exists specifically to stop that "bleeding" and collapse the fragments into one locus. Distance
clumping has no way to know two windows tag the same causal variant, so it can over-count. It is the
honest choice only when no matched genotype panel is available to compute pairwise r² — which is the
case here.*

# Fine-mapping + coding annotation (Step 4)

*SuSiE fine-mapping per locus (MHC excluded), then missense/splice annotation of credible-set
variants to nominate a candidate effector gene. **Caveat:** out-of-sample LD may miscalibrate
credible sets — interpret set sizes and PIPs with that in mind.*

```{r finemap-table}
annot_tab <- read_many(sm@input$annot)
show_table(annot_tab, "Credible-set variants and coding annotation")
```

Only the credible-set variants annotated as **missense** are shown below — these are the coding
candidates that fine-mapping nominates as potentially causal. The recovery of textbook LDL coding
variants (PCSK9 R46L, APOE C130R) at PIP = 1 is the strongest validation that the pipeline works
despite the out-of-sample LD.

```{r finemap-missense}
# Highlight the coding candidates. annot.tsv carries every credible-set variant; the missense
# subset is the biologically interpretable headline.
if (!is.null(annot_tab) && nrow(annot_tab) && "is_missense" %in% names(annot_tab)) {
  keep <- intersect(c("gene_symbol","ID","aa_change","pip","cs","sift_pred",
                      "polyphen_pred","is_known_causal"), names(annot_tab))
  miss <- annot_tab[is_missense == TRUE, ..keep][order(-pip)]
  show_table(miss, "Missense credible-set variants (coding effector candidates)")
} else {
  show_table(NULL)
}
```

```{r finemap-fig, results='asis'}
# Two cross-locus summary figures (per-locus QC counts; coding-vs-regulatory consequence mix).
show_figs(c(sm@input$finemap_summary, sm@input$consequence_summary), cols = 1)
```

```{r finemap-regional-fig, results='asis'}
# Per-locus LocusZoom-style regional plots (GWAS -log10P coloured by r2-to-lead over a PIP track).
regional_pngs <- unlist(lapply(sm@input$finemap_plots,
                               function(d) list.files(d, "\\.png$", full.names = TRUE)))
show_figs(regional_pngs)
```

## MCQ — Fine-mapping

**The pipeline fine-maps with an out-of-sample 1000G EUR LD reference rather than the in-sample
UKBB LD used by the reference paper. The primary risk this introduces to the credible sets is:**

- A. They will always be too small.
- B. LD mismatch can miscalibrate posterior inclusion probabilities and yield spurious or oversized credible sets.
- C. SuSiE cannot run at all without in-sample LD.
- D. It only affects the Manhattan plot, not fine-mapping.

*Answer: **B.** SuSiE-RSS assumes the LD matrix matches the GWAS sample; a mismatched reference
distorts PIPs and credible-set membership. In this run the fingerprint is visible in the
convergence QC (most loci hit the IBSS iteration cap), which is why an `estimate_s_rss` LD-mismatch
diagnostic is reported per locus and `estimate_residual_variance` is fixed to `FALSE` — the
documented remedy for out-of-sample LD. This is the single most important caveat of the analysis.*

# Limitations

- **Out-of-sample LD.** Fine-mapping and distance-clumping use a 1000G EUR reference rather than
  in-sample LD; SuSiE credible sets may be miscalibrated (flagged per locus by the kriging
  diagnostic). This is the single most important caveat and the main methodological divergence
  from the reference paper, which had in-sample UKBB LD.
- **Distance-only locus definition.** With no genotype panel provided at Step 3, independent loci
  are defined by a ±1 Mb distance rule rather than the paper's r²≥0.05 LD merge; a very strong
  signal can bleed past the window and be split into apparent multiple loci (see the Step-3 MCQ).
- **Fine-mapping scoped to LDL effector loci.** Step 4 fine-maps the 11 canonical LDL effector
  loci (a hypothesis-driven subset), not genome-wide — chosen so out-of-sample-LD behaviour can be
  inspected locus by locus.
- **Coding annotation explains few loci.** Of the credible-set variants, only a handful are
  missense; most signal is regulatory/intronic. Resolving those mechanisms needs molecular-QTL
  colocalization — the next step, not yet included here.
- **EUR-only.** Exposures and (planned) outcomes are European-ancestry, so estimates may not
  generalize — a deliberate choice to keep LD and allele frequencies matched.

*Steps 5–7 (colocalization, Mendelian randomization) are in progress and will be added to this
report, each with its own MCQ, once their pipeline rules have run.*

